\newcommand{\mrfNairInferenceResultsCaption}{Comparison of decoding quality and runtime across SC and MIS instances, showing average relative optimality gaps and total runtimes for canonical MRFs, learned MRFs, and neural baselines.}
\newcommand{\mrfNairInferenceGroupedBarOptimalityCaption}{Average relative optimality gap across SC and MIS instances for canonical MRFs, learned MRFs, and neural baselines under EGG and GD decoding.}
\newcommand{\mrfNairInferenceStackedGroupedBarRuntimeCaption}{Runtime decomposition across SC and MIS instances, showing the contribution of inference and decoding time for canonical MRFs, learned MRFs, and neural baselines under EGG and GD decoding.}
\newcommand{\inferenceThetaDistributionCaption}{Distribution of predicted local interaction parameters for LMRF-LBP and LMRF-NMF. Predicted non-unary factor potentials are grouped according to whether the corresponding local assignment is feasible or infeasible.}
\newcommand{\bpgdDeltaMarginalsCaption}{Average per-step change in marginals when clamping evidence during GD with LBP marinal inference. Positive values mean the marginal decreased after the update (since $\Delta=\text{before}-\text{after}$). Curves are averaged over the test set.}
\newcommand{\inferenceFreqMaxIterAblationParetoCaption}{Runtime--quality Pareto frontiers for GD ablations. The left column varies the reinference interval with fixed message-passing depth, while the right column varies the maximum number of inference iterations with fixed reinference interval. Points are annotated by the corresponding hyperparameter value.}
\newcommand{\inferenceTestNairGapBarCaption}{Optimality gaps for fix-and-solve with learned predictions. For each fixed-variable proportion, the decoder fixes the corresponding fraction of variables, after which SCIP solves the reduced problem over the remaining variables under a $300$s time limit. Gaps are measured relative to the Gurobi optimum of the original instance.}
\newcommand{\inferenceTestNairRuntimeStackedBarCaption}{Runtime decomposition for fix-and-solve with learned predictions. Total runtime is split into preprocessing, model forward pass, decoder-based fixing, and the final SCIP solve on the reduced problem under a $300$s time limit.}
\newcommand{\inferenceTestNairParetoCaption}{Runtime--quality Pareto frontiers for fix-and-solve with learned predictions. Each point corresponds to one fixed-variable proportion, with lower and further-left points indicating lower optimality gap and runtime. Gaps are measured relative to the Gurobi optimum, while SCIP solves each reduced problem under a $300$s time limit.}