\chapter*{Acknowledgements}
    
This thesis has been a long-awaited achievement, and I would like to take the time to thank everyone who contributed, in one way or another, to making it possible.

First and foremost, I would like to express my sincere gratitude to my supervisors, Roberto Wolfler-Calvo, Mathieu Lacroix, and Joseph Le Roux, for their guidance, support, and involvement throughout these four years of work. I learned a great deal from each of them. Mathieu’s rigorous analysis of my work, sometimes leading to lively discussions during our meetings, helped me remain precise, critical, and focused. I am grateful for his directness and for the high standards he encouraged me to maintain. Joseph’s insights into methodology were a major driving force behind the machine learning aspects of this thesis, and his guidance played an essential role in shaping its final direction. Roberto provided constant support and encouragement throughout the thesis, as well as valuable perspective on combinatorial optimization and its applications. I am also deeply thankful for the moral support and guidance he offered along the way.

I would also like to thank the LIPN and the wider Université Sorbonne Paris Nord community for providing such a welcoming, stimulating, and dynamic research environment. My thanks go to the past and current PhD students with whom I shared this experience, including Alexandre Dupont-Bouillard, Daria Pchelina, Mathieu Vallée, Yue Zhang, Alexis Schneider, Jules Nicolas-Thouvenin, Alexandre Xia, Francesca Demelas, Baptiste Deshayes, Carole Porrier, and Frédéric Jehan. Their presence, conversations, and good humour helped create an animated and supportive working atmosphere.

I am also grateful to Sébastien Martin and Huawei Technologies. Sébastien was my initial point of contact in finding this PhD opportunity following my internship at the Huawei offices in Boulogne-Billancourt. The collaborative project between Huawei and the AOC group at LIPN, on multi-commodity flow column generation solvers, provided the initial framework and funding for my doctoral contract.

Finally, I would like to thank my friends and family, who supported me throughout these four years. They listened patiently while I tried to explain my work, offered their opinions on my writing, and provided encouragement and support when I needed it most. Without them, completing this thesis would have been much more difficult.

\thesisclearpage