\chapter{Introduction}\label{chap0:intro}
\chaptermark{Introduction}

\section{Motivation}

Many decision-making problems in computer science, operations research, artificial intelligence, and engineering can be formulated as combinatorial  optimization problems. These problems require selecting, assigning, ordering, or grouping discrete objects in order to optimize a given objective while satisfying a set of constraints. Examples include selecting a minimum set of resources to cover a collection of requirements, finding a maximum subset of mutually compatible decisions, scheduling jobs under resource limitations, or planning routes through a network. Although their applications are diverse, these problems share a common mathematical difficulty: the feasible set is discrete and often grows exponentially with the size of the instance.

Integer Linear Programming (ILP) provides a general and expressive framework for representing such problems. In particular, Binary Integer Linear Programs (BILPs), whose variables take values in {0,1}, can model a wide range of classical combinatorial optimization problems. In this thesis, BILPs are not only treated as algebraic optimization problems, but also as structured prediction problems. A feasible solution is a binary assignment, but the value of each variable cannot generally be predicted independently: feasibility constraints couple variables together, and the quality of a solution depends on the global configuration. This coupling is the central challenge addressed throughout the thesis.

Modern ILP solvers combine many algorithmic components, including branch-and-bound, cutting planes, primal heuristics, rounding procedures, and local improvement methods. These techniques are highly effective, but they may still be expensive on large or difficult instances. Moreover, many solver decisions are heuristic in nature: which variable to branch on, which node to explore, which partial assignment to pursue, or which variables to fix. This has motivated a growing literature on learning-augmented optimization, where machine learning models are trained to predict useful information from previously solved instances.

Graph neural networks have become a natural tool in this setting because ILP instances can be represented as variable-constraint bipartite graphs. A neural model can process this graph and produce scores for decision variables, constraints, or solver actions. However, a common limitation of variable-level neural prediction is that the final predictive distribution is often factorized across variables. The graph encoder may propagate constraint information into the embeddings, but the output itself typically assigns a probability or score to each variable independently. This can lead to assignments that are locally plausible but globally inconsistent.

This thesis explores a complementary direction. Instead of treating learned predictions as independent variable scores, it uses Markov Random Fields (MRFs) and approximate inference to represent dependencies among decision variables explicitly. In this view, constraints induce factors over groups of variables, learned or canonical log-potentials encode local compatibility, and inference algorithms such as Loopy Belief Propagation (LBP) or Naive Mean-Field (NMF) transform local information into structured marginal beliefs. These beliefs can then guide solution construction, variable fixing, and refinement.

The guiding idea is that learning should not only predict which variables are likely to be active in a good solution, but should also learn how such predictions interact through the constraint structure of the optimization problem.

\section{Research Problem}

The main research question addressed in this thesis is: \emph{How can machine learning and graphical-model inference be combined to produce structured, constraint-aware predictions for binary integer linear programs?}

This question is motivated by three observations.

First, BILP feasibility constraints create dependencies between variables. A prediction method that treats variables independently may ignore important local compatibility conditions. For example, in Maximum Independent Set, selecting one vertex immediately restricts the selection of its neighbors. In Set Cover, selecting or not selecting a set affects which coverage constraints remain satisfied. A useful predictor must therefore account for the consequences of partial assignments.

Second, graphical models provide a natural language for representing such dependencies. MRFs define distributions over joint assignments through local factors. If BILP constraints are mapped to factors, inference in the resulting graphical model can propagate information between variables that participate in common constraints. This suggests that approximate inference can be used as a structured layer between neural prediction and discrete decoding.

Third, exact inference in graphical models is generally intractable, and direct factor-table representations become expensive when constraints involve many variables. Therefore, a practical learning-based approach must balance expressiveness, scalability, and decoding quality. This thesis studies several points along this trade-off: direct MRF inference with learned factors, iterative refinement dynamics augmented with MRF information, and approximate low-arity factor representations for scalable inference.

\section{Thesis Approach}

The thesis develops a sequence of models that connect BILP formulations, neural graph encoders, MRF parameterizations, and approximate inference.

The starting point is the variable-constraint graph of a BILP. Variable nodes represent decision variables, constraint nodes represent linear constraints, and edges indicate nonzero coefficients. This graph provides the input structure for neural encoders. The same BILP constraints also motivate a factor graph over solution variables, where each factor represents a local interaction among variables participating in a constraint.

The first contribution is to replace independent Bernoulli variable prediction with an exponential-family MRF over BILP variables. In this model, constraints define factor scopes and a graph neural network predicts factor log-potentials. Fixed-iteration LBP or NMF then produces approximate marginals. These marginals are used both as training signals and as decoding signals. In particular, guided decimation constructs solutions by progressively fixing variables and recomputing beliefs under the accumulated partial assignment.

The second contribution extends the inference view toward learned refinement dynamics. Instead of producing a solution in a single pass, the model iteratively transforms a noisy, relaxed, or partially specified assignment toward a high-quality discrete solution. This connects MRF-based structured prediction with denoising and flow-matching ideas: solution generation is treated as a trajectory rather than a one-shot decision. MRF-derived information is used to augment the state representation, allowing the refinement model to exploit both neural features and structured inference signals.

The third contribution addresses scalability. Direct over-complete MRF factor tables grow exponentially with factor arity, which is problematic for constraints involving many variables, such as Set Cover constraints. To reduce this cost, the thesis studies feature-to-factor mappings that project information from the original high-arity structure into approximate lower-arity factor graphs. These mappings preserve some dependency information while making inference more tractable. Several mapping variants are considered, including permutation-invariant aggregation, variable-constraint feature separation, attention-based aggregation, and transformer-based mappings.

Together, these contributions support a unified perspective: learned optimization methods can benefit from treating BILP solution prediction as approximate inference in a structured probabilistic model, rather than as independent variable classification.

\section{Contributions}

The main contributions of this thesis are the following.

\begin{enumerate}
    \item \textbf{A structured MRF formulation for BILP solution prediction.} The thesis formulates binary solution prediction as inference in an exponential-family MRF whose factor structure is derived from the constraints of a BILP. This provides an explicit probabilistic model of dependencies between decision variables.
    \item \textbf{Canonical and learned MRF parameterizations.} A canonical feasibility-cost MRF is introduced to encode objective terms and constraint feasibility directly as log-potentials. This is extended to learned MRFs, where a graph neural network predicts over-complete factor parameters from the variable-–constraint graph.
    \item \textbf{Differentiable approximate inference for training and decoding.} Fixed-iteration LBP and NMF are used as differentiable inference layers. Their approximate marginals provide structured prediction signals that can be used in training objectives and in solution construction.
    \item \textbf{Inference-guided decoding through guided decimation.} The thesis studies decoding strategies that use inferred marginals to construct feasible solutions. Guided decimation repeatedly conditions the MRF on partial assignments and recomputes beliefs, providing an adaptive form of inference-guided solution construction.
    \item \textbf{Learned refinement dynamics for solution generation.} The thesis connects structured inference with denoising and flow-based refinement. Instead of relying only on a single prediction pass, learned dynamics iteratively update candidate assignments toward high-quality solutions.
    \item \textbf{Approximate feature-to-factor mappings for scalable inference.} To address the cost of high-arity factor tables, the thesis proposes approximate lower-arity factor representations and studies several neural mappings from graph features to factor parameters.
    \item \textbf{Empirical evaluation on representative BILP families.} The proposed methods are evaluated on Maximum Independent Set and Set Cover instances, which expose complementary structures: pairwise exclusion constraints and higher-arity coverage constraints.
\end{enumerate}

\section{Organization of the Thesis}

The remainder of the thesis is organized as follows.

Chapter~\ref{chap1:background} introduces the background required for the thesis. It reviews integer linear programming, binary formulations, branch-and-bound and primal heuristics, graph representations of ILPs, machine learning concepts, graph neural networks, Markov Random Fields, and approximate inference.

Chapter~\ref{chap:chapter_2_related_work} reviews related work. It positions the thesis with respect to learning-augmented ILP solvers, neural solution prediction, graphical-model approaches to optimization, and iterative generative solvers such as denoising and flow-based methods.

Chapter~\ref{chapter:3_differentiable_mrf_inference_for_BILPs} presents differentiable MRF inference for BILPs. It introduces canonical and learned MRF parameterizations, uses LBP and NMF as approximate inference layers, and studies decoding strategies such as entropy-guided greedy decoding and guided decimation.

Chapter~\ref{chapter:4_denoising_and_mrf_inference} moves from structured inference to learned refinement dynamics. It studies MRF-augmented flow and denoising models that iteratively transform initial candidate assignments into improved solutions.

Chapter~\ref{chapter:5_feature-to-factor_mappings_for_approximate_mrf_inference} investigates feature-to-factor mappings for approximate MRF inference. It addresses the scalability limitations of high-arity factor tables by constructing lower-arity approximate factor graphs and learning mappings from graph embeddings to factor parameters.

The final chapter concludes the thesis by summarizing the main findings, discussing limitations, and outlining directions for future research.
