\section{Machine Learning and Deep Learning}
\label{sec:ml_intro}

Machine Learning (ML), and in particular \emph{Deep Learning} (DL), provides a framework for learning parameterized models from data rather than designing all decision rules by hand~\cite{lecun_deep_2015,goodfellow_deep_2016}. In this thesis, ML is used to learn mappings from structured optimization instances to quantities that help construct high-quality solutions: variable scores, MRF parameters, approximate marginals, or denoising directions.

In general, a machine learning model is a parameterized function 
\begin{equation}
    f_{\GENENNPARAM} : \mathcal{X} \rightarrow \mathcal{Y},
\end{equation}
where $\mathcal{X}$ denotes the input domain, $\mathcal{Y}$ the output domain, and $\GENENNPARAM \in \GENENNPARAMSPACE$ denotes the trainable parameters. The model is evaluated according to a task-specific loss function $\mathcal{L}$ defined on training examples. The goal of training is to find parameters $\GENENNPARAM^*$ that minimize the expected loss,
\begin{equation}
    \label{equ:training_task}
    \GENENNPARAM^*
    =
    \argmin_{\GENENNPARAM \in \GENENNPARAMSPACE}
    \mathbb{E}_{z \sim \DATADIST}
    \left[
        \mathcal{L}(\GENENNPARAM,z)
    \right],
\end{equation}
where $\DATADIST$ denotes the data distribution and $z$ denotes a generic training example.

The learning problems considered later differ in their outputs and losses. Some models are trained in a supervised way from reference solutions; others use denoising or flow-matching objectives to generate candidate solutions through iterative refinement. The common theme is that the output is not an isolated label, but a structured prediction signal tied to the variables and constraints of a BILP.


\subsection{Supervised Learning}
\label{sec:supervised_learning}

Supervised learning refers to the class of learning strategies in which the \emph{ground truth} or target value is available during training. Formally, we assume access to a dataset of input–output pairs 
\begin{equation}
    \DATASET = \{(x_i, y_i)\}_{i=1}^{N}, \quad x_i \in \mathcal{X}, \; y_i \in \mathcal{Y},
\end{equation}
where each input $x_i$ is associated with a corresponding label $y_i$ representing the correct or desired outcome. The objective is to learn a parameterized function $f_{\GENENNPARAM} : \mathcal{X} \rightarrow \mathcal{Y}$ capable of producing predictions $\hat{y} = f_{\GENENNPARAM}(x)$ that generalize well to unseen data drawn from the same underlying distribution $\DATADIST$.

By quantifying the discrepancy between the predicted outcome $\hat{y}$ and the ground truth $y$, it becomes possible to define a \emph{loss function} $\SUPLOSS : \GENENNPARAMSPACE \times \mathcal{X} \times \mathcal{Y} \rightarrow \mathbb{R}$ that measures model performance on each training example. The learning problem is then formulated as minimizing the expected loss over the data distribution:
\begin{equation}
    \label{equ:supervised_learning}
    \GENENNPARAM^* = \argmin_{\GENENNPARAM \in \GENENNPARAMSPACE} \, \mathbb{E}_{(x,y) \sim \DATADIST} \left[ \SUPLOSS(\GENENNPARAM, x, y) \right].
\end{equation}

In practice, this expectation is approximated by the empirical average over the training set:
\begin{equation}
    \widehat{\mathcal{L}}_{\mathrm{sup}}(\GENENNPARAM) = \frac{1}{N} \sum_{i=1}^{N} \SUPLOSS(\GENENNPARAM, x_i, y_i).
\end{equation}

The form of $\SUPLOSS$ depends on the prediction task:
\begin{itemize}
    \item In \emph{regression} tasks, the model predicts continuous quantities and typical losses include the squared error $\SUPLOSS = \| f_{\GENENNPARAM}(x) - y \|^2$ or the mean absolute error.
    \item In \emph{classification} tasks, where the label $y$ belongs to a discrete set of classes, the loss commonly takes the form of the negative log-likelihood or cross-entropy:
    \begin{equation}
        \SUPLOSS(\GENENNPARAM, x, y) = - \log p_{\GENENNPARAM}(y|x),
    \end{equation}
    where $p_{\GENENNPARAM}(y|x)$ denotes the model’s predicted probability for the correct class.
\end{itemize}

For binary prediction tasks, where $y\in\{0,1\}$ and the model outputs a probability $p_{\GENENNPARAM}(x)\in(0,1)$, the corresponding binary cross-entropy is
\begin{equation}
    \BCELOSS(\GENENNPARAM,x,y)
    =
    -y\log p_{\GENENNPARAM}(x)
    -(1-y)\log\!\left(1-p_{\GENENNPARAM}(x)\right).
\end{equation}
This loss is used for direct variable-level predictors, including the factorized GNN baselines in Chapter~\ref{chapter:3_differentiable_mrf_inference_for_BILPs}.

Supervised learning thus reduces to a parameter optimization problem, where $\GENENNPARAM$ is iteratively updated to minimize $\widehat{\mathcal{L}}_{\mathrm{sup}}(\GENENNPARAM)$ using stochastic gradient descent or one of its many variants \cite{goodfellow_deep_2016,kingma_adam_2014}. When the learned parameters generalize beyond the training data, the model effectively captures an underlying mapping between inputs and outputs, a property central to most modern applications of deep learning \cite{vapnik_nature_2000,bishop_pattern_2006}.

\subsection{Generative Models and Denoising Objectives}
\label{sec:generative_denoising}

The supervised setting described above assumes access to reference targets and trains a model to reproduce them. In combinatorial optimization, however, reference solutions can be expensive to obtain, and a single instance may admit many optimal or near-optimal solutions. This motivates generative and denoising-based approaches, which learn a distribution or refinement process over candidate assignments rather than a single deterministic mapping.

Generative models aim to learn a probability distribution over the data, rather than directly predicting a single deterministic output. In the conditional setting considered throughout this thesis, an input instance $x \in \mathcal{X}$ may be associated with multiple valid or high-quality outputs $y \in \mathcal{Y}$. This is particularly relevant in combinatorial optimization, where a problem instance can admit several near-optimal solutions with different structures. Instead of returning a single prediction, a generative model defines a conditional distribution or generative process over possible outputs and obtains predictions by sampling, integrating learned dynamics, or selecting high-probability candidates (see Figure~\ref{fig:conditional_generative_graph}).

Formally, we view input-output pairs as realizations of random variables $X$ and $Y$. The goal is to approximate the conditional distribution
\begin{equation}
    p(y \mid x) = \mathbb{P}\left[ Y = y \mid X = x \right],
\end{equation}
where $x$ denotes the observed input instance and $y$ denotes a target output, such as a feasible solution or an optimal assignment.

In practice, the true conditional distribution $p(y \mid x)$ is unknown and is approximated by a parameterized model $p_{\GENENNPARAM}(y \mid x)$. In some architectures, a neural network first maps the input instance $x$ to output-distribution parameters $\eta=f_{\GENENNPARAM}(x)$, yielding a distribution $p_{\eta}(y)$. Predictions can then be obtained by sampling
\begin{equation}
    y \sim p_{\GENENNPARAM}(\cdot \mid x),
\end{equation}
or by selecting high-probability outputs according to the learned distribution.

Different families of generative models mainly differ in how they parameterize this distribution and in the training objective used to match it to the data distribution. Examples include likelihood-based objectives, denoising objectives, diffusion-style objectives, and flow-matching objectives. In the remainder of this section, we focus on denoising-based generative models, since they provide the conceptual basis for the diffusion and flow-based refinement methods studied in Chapter~\ref{chapter:4_denoising_and_mrf_inference}.

\input{resources/tikz/chapter_1/ml/generative_model.tex}

\paragraph{Denoising-based generative modeling}

A central idea in modern generative modeling is to learn how to recover structured data from corrupted observations. Instead of directly learning to sample from the data distribution, denoising-based methods define a corruption process that progressively perturbs a clean target $y$ into a noisy version $\tilde{y}$. A model is then trained to invert this process, either by predicting the original clean sample, the noise that was added, or an update direction that moves the corrupted sample closer to the data distribution.

Let $y \in \mathcal{Y}$ denote a clean target output, such as an optimal or high-quality solution of a combinatorial optimization problem. A corruption process defines a conditional distribution
\begin{equation}
    q(\tilde{y} \mid y, t),
\end{equation}
where $t$ denotes the noise level or time index. For small values of $t$, the corrupted sample $\tilde{y}$ remains close to the original target, while for large values of $t$ it may contain little information about $y$. The learning problem consists in training a model that, given the corrupted sample $\tilde{y}$, the input instance $x$, and the noise level $t$, predicts how to recover information about the clean output:
\begin{equation}
    \DENOISER(x, \tilde{y}, t) \approx y.
\end{equation}
Equivalently, depending on the formulation, the model may predict the perturbation applied to $y$, or a direction in the output space that transforms $\tilde{y}$ toward a clean sample.

Figure~\ref{fig:graph_denoising_illustration} illustrates this view for graph-structured binary assignments.

\input{resources/tikz/chapter_1/ml/graph_denoising_illustration.tex}

This denoising view is especially natural for structured prediction. In many problems, the output is not a single scalar label but a structured object composed of many dependent decisions. In combinatorial optimization, for example, a solution may be represented by a binary vector $y \in \{0,1\}^n$, where each component corresponds to the value of a decision variable. Corrupting such a solution can be interpreted as randomly masking, flipping, or relaxing some of its variables. The denoising model must then learn how local and global dependencies between variables constrain the reconstruction of a valid or high-quality solution.

A common training objective is to minimize a reconstruction loss between the model prediction and the clean target. For example, when the model predicts the original solution, one may use
\begin{equation}
    \DENOISELOSS(\GENENNPARAM)
    =
    \mathbb{E}_{(x,y)\sim\DATADIST,\, t,\, \tilde{y}\sim q(\cdot\mid y,t)}
    \left[
        \ell\left(\DENOISER(x,\tilde{y},t), y\right)
    \right],
\end{equation}
where $t$ is sampled from a chosen noise-level distribution and $\ell$ is a suitable loss function, such as a cross-entropy loss for discrete variables or a squared error loss for continuous relaxations. The model is therefore trained on many reconstruction tasks of varying difficulty, ranging from lightly corrupted samples to highly noisy ones.

At inference time, denoising-based generative models reverse the corruption process. Starting from a noisy or partially initialized sample, the learned denoising model is applied iteratively to produce increasingly structured outputs. This procedure can be interpreted as a learned refinement process: each denoising step uses the instance information $x$, the current candidate solution, and the noise level to move the sample toward the learned solution distribution.

In the context of this thesis, this perspective is useful because it connects generative modeling with solution construction. Rather than predicting each variable independently in a single forward pass, a denoising model defines iterative dynamics over candidate solutions. This makes it possible to progressively refine a noisy assignment into a feasible or high-quality solution, while potentially capturing dependencies between variables through the architecture used to parameterize the denoising function.

This general principle underlies both diffusion models, which learn to reverse a stochastic noising process, and flow-based formulations, which learn a deterministic or approximately deterministic transformation from a simple initial distribution toward the data distribution.

\paragraph{Flow matching and velocity prediction models}
\label{par:background_flow_matching}

Flow matching is a simulation-free training framework for continuous normalizing flows. This presentation follows the flow matching framework of Lipman et al.~\cite{lipman_flow_2022} and the conditional flow matching formulation of Tong et al.~\cite{kerrigan_dynamic_2024}. The goal is to learn a time-dependent vector field that transports a simple source distribution $p_0$ to a target data distribution $p_1=\DATADIST$. Let $p_t$ denote a probability path on the solution space, with $t\in[0,1]$, satisfying $p_{t=0}=p_0$ and $p_{t=1}=p_1$. A vector field $u_t$ is said to generate this path if it satisfies the continuity equation
\begin{equation}
    \partial_t p_t(y) + \nabla_y \cdot \bigl(p_t(y) u_t(y)\bigr) = 0.
\end{equation}
Flow matching trains a parameterized vector field $v_{\GENENNPARAM}(y,t)$ to match such a target vector field along the chosen probability path.

In practice, the marginal path $p_t$ and its corresponding marginal vector field $u_t$ are not directly available from finite data. Flow matching therefore uses conditional probability paths. Given a data sample $y_1 \sim \DATADIST$, one defines a conditional path $p_t(y\mid y_1)$ that starts from a simple distribution at $t=0$ and concentrates around $y_1$ at $t=1$, together with a conditional vector field $u_t(y\mid y_1)$ that generates this conditional path. The marginal path is then
\begin{equation}
    p_t(y) = \int p_t(y\mid y_1) \, \DATADIST(y_1) \, dy_1.
\end{equation}
The corresponding marginal vector field is obtained by averaging the conditional vector fields under the posterior over conditioning samples,
\begin{equation}
    u_t(y)
    =
    \mathbb{E}_{y_1 \sim \DATADIST}
    \left[
        u_t(y\mid y_1)
        \frac{p_t(y\mid y_1)}{p_t(y)}
    \right].
\end{equation}
The key result behind conditional flow matching is that one can train using samples from the conditional paths without explicitly computing this marginal vector field.

A common choice is an optimal-transport-style conditional path obtained by pairing a source sample $y_0\sim p_0$ with a target sample $y_1\sim\DATADIST$ and setting
\begin{equation}
    y_t = (1-t)y_0 + t y_1.
\end{equation}
For this particular path, the conditional target velocity is
\begin{equation}
    u_t(y_t \mid y_0,y_1) = y_1-y_0.
\end{equation}
This straight-line construction should be understood as one possible conditional path, not as the definition of flow matching itself.

In a conditional generation setting, the target distribution depends on an input instance $x$. For example, in combinatorial optimization, $x$ may represent the optimization instance and $y_1$ may represent an optimal or high-quality solution. The model is then written as $v_{\GENENNPARAM}(y,t,x)$, and the conditional flow-matching objective becomes
\begin{equation}
    \FMLOSS(\GENENNPARAM)
    =
    \mathbb{E}_{(x,y_1)\sim\DATADIST,\; y_t\sim p_t(\cdot\mid y_1),\; t\sim U[0,1]}
    \left[
        \left\|
            v_{\GENENNPARAM}(y_t,t,x) - u_t(y_t\mid y_1)
        \right\|_2^2
    \right].
\end{equation}
For the straight-line path above, this can equivalently be written by sampling $y_0\sim p_0$, setting $y_t=(1-t)y_0+ty_1$, and using the target velocity $u_t=y_1-y_0$:
\begin{equation}
    \FMLOSS(\GENENNPARAM)
    =
    \mathbb{E}_{(x,y_1)\sim\DATADIST,\; y_0\sim p_0,\; t\sim U[0,1]}
    \left[
        \left\|
            v_{\GENENNPARAM}(y_t,t,x) - (y_1-y_0)
        \right\|_2^2
    \right].
\end{equation}

Once the velocity field has been learned, samples are generated by drawing an initial state $y(0)\sim p_0$ and solving the ordinary differential equation
\begin{equation}
    \frac{d y(t)}{dt}
    =
    v_{\GENENNPARAM}(y(t),t,x),
    \qquad
    y(0)\sim p_0.
\end{equation}
The final state $y(1)$ is then decoded, if necessary, into a feasible discrete solution. When the target output is discrete, as in BILP solution generation, the flow is defined on a continuous relaxation or embedding of the solution space, and a decoding or rounding step is required to obtain a discrete assignment. Algorithm~\ref{alg:flow_sampling} summarizes the resulting sampling procedure using a simple explicit Euler discretization of the learned flow ODE.

\begin{algorithm}[t]
\caption{Conditional generation with a learned flow}
\label{alg:flow_sampling}
\DontPrintSemicolon

\KwIn{
    Problem instance $x$;
    learned vector field $v_{\GENENNPARAM}$;
    source distribution $p_0$;
    number of integration steps $K$;
    decoding operator $\mathrm{Dec}$
}
\KwOut{
    Discrete candidate solution $\hat{y}$
}

Sample an initial state $y^{(0)} \sim p_0$\;
Set $\Delta t \leftarrow 1/K$\;

\For{$k = 0,\ldots,K-1$}{
    Set $t_k \leftarrow k \Delta t$\;
    
    Evaluate the vector field $\hat{u}^{(k)} \leftarrow v_{\GENENNPARAM}(y^{(k)}, t_k, x)$\;
    
    Update the state using an Euler step:
    $y^{(k+1)} \leftarrow y^{(k)} + \Delta t \, \hat{u}^{(k)}$\;
}

Decode the final state $\hat{y} \leftarrow \mathrm{Dec}(y^{(K)}, x)$\;

\Return{$\hat{y}$}\;

\end{algorithm}

\paragraph{From data generation to solution generation}
In standard generative modeling, the objective is to sample realistic data points, such as images or signals. In this thesis, the generated object is instead a candidate solution to an optimization instance. This introduces an additional requirement: intermediate states of a denoising or flow model may be noisy, continuous, or infeasible, whereas the final output must be decoded into a discrete assignment satisfying the constraints of the original BILP. we focus on denoising-based generative models, since they provide the conceptual basis for the diffusion and flow-based refinement methods studied in Chapter~\ref{chapter:4_denoising_and_mrf_inference}.


\subsection{Neural Networks and Graph Neural Networks}
\label{sec:neural_networks}

Neural networks are parametric models designed to approximate complex nonlinear mappings between inputs and outputs \cite{hornik_multilayer_1989,goodfellow_deep_2016}. A neural network defines a function 
\begin{equation}
    f_{\GENENNPARAM} : \mathcal{X} \rightarrow \mathcal{Y},
\end{equation}
composed of a sequence of linear transformations followed by nonlinear activation functions:
\begin{equation}
    h^{(l)} = \sigma\!\left(W^{(l)} h^{(l-1)} + b^{(l)}\right), \quad l = 1, \dots, L,
\end{equation}
where $h^{(0)} = x \in \mathcal{X}$ is the input, $h^{(L)}$ represents the output layer, and the trainable parameters $\GENENNPARAM$ collect the weights and biases $\{W^{(l)}, b^{(l)}\}_{l=1}^{L}$. The activation function $\sigma$ introduces nonlinearity (e.g., ReLU, sigmoid, or tanh), which allows the network to model complex relationships that cannot be captured by purely linear models. A feed-forward network of this form, composed of several affine layers and nonlinear activations, is often called a Multi-Layer Perceptron (MLP).

During training, the parameters $\GENENNPARAM$ are optimized to minimize a task-specific loss function (see Equation~\ref{equ:training_task}), typically using stochastic gradient-based methods such as backpropagation and variants of gradient descent \cite{rumelhart_learning_1986,kingma_adam_2014}. The success of neural networks in computer vision, language modeling, and reinforcement learning has motivated their extension to domains where the data exhibit rich, relational or graph-structured dependencies.

\input{resources/tikz/chapter_1/ml/nn_tikz}

\paragraph{Graph Neural Networks}
Many real-world problems can be naturally represented as graphs $\mathcal{G} = (\mathcal{V}, \mathcal{E})$, where $\mathcal{V}$ is the set of nodes and $\mathcal{E}$ the set of edges, encoding pairwise relationships or constraints. Traditional neural architectures (e.g., CNNs, RNNs) assume Euclidean input domains and therefore fail to directly exploit the topological structure of such data. \emph{Graph Neural Networks} (GNNs) address this limitation by defining message-passing operations that iteratively exchange information between neighboring nodes \cite{scarselli_graph_2009,kipf_semi-supervised_2017,bronstein_geometric_2017}. 

Each node $i \in \mathcal{V}$ is associated with an initial feature vector $h_i^{(0)} \in \mathbb{R}^{d_0}$, and the node representations are updated through successive \emph{message passing} layers of the general form:
\begin{equation}
    h_i^{(l)} = \phi^{(l)} \!\left( h_i^{(l-1)}, \, \bigoplus_{j \in \mathcal{N}(i)} \psi^{(l)}(h_i^{(l-1)}, h_j^{(l-1)}, e_{ij}) \right),
    \label{eq:gnn_message_passing}
\end{equation}
where $\mathcal{N}(i)$ denotes the neighborhood of node $i$, $e_{ij}$ represents optional edge features, $\bigoplus$ is a permutation-invariant aggregation operator (e.g., sum, mean, or max), and $\psi^{(l)}$ and $\phi^{(l)}$ are neural functions parameterized by trainable weights. This formulation is permutation-equivariant at the node level: permuting the node order permutes the output representations in the same way. Graph-level readouts are typically designed to be permutation-invariant.

After several message-passing iterations, the node representations can be used directly for node-level predictions, combined pairwise for edge-level predictions, or pooled through a readout function for graph-level predictions.

GNNs are especially useful for structured prediction problems in which the input is naturally relational or graph-structured~\cite{gilmer_neural_2017,battaglia_relational_2018}. In combinatorial optimization, they provide a way to learn heuristics or prediction rules that operate directly on graph-structured problem instances.

In the context of this thesis, GNNs provide one of the main neural representation mechanisms for BILP instances. They process the variable-constraint graph and produce embeddings that can be used for direct variable-level prediction, MRF factor-parameter prediction, denoising updates, or as inputs to subsequent feature-to-factor mapping modules. This distinction motivates the feasibility-preserving decoding procedures in Section~\ref{subsubsec:chapter_3_egg_gd} and the refinement-based decoding procedures in Section~\ref{sec:chapter_4_unrolling_decoding}.

\paragraph{Transformers and attention}

Transformers are neural architectures based on self-attention, originally introduced for sequence modeling~\cite{vaswani_attention_2017}. Given a collection of $n$ input embeddings of dimension $d$, arranged as rows of a matrix $\TRFH\in\mathbb{R}^{n\times d}$, a self-attention layer first computes queries, keys, and values
\begin{equation}
    \TRFQ = \TRFH \TRFWQ,
    \qquad
    \TRFK = \TRFH \TRFWK,
    \qquad
    \TRFV = \TRFH \TRFWV,
\end{equation}
where $\TRFWQ\in\mathbb{R}^{d\times d_k}$, $\TRFWK\in\mathbb{R}^{d\times d_k}$, and $\TRFWV\in\mathbb{R}^{d\times d_v}$ are learned projection matrices. The quantity $d_k$ denotes the key/query dimension and is used to scale the dot products. The representations are then updated through scaled dot-product attention:
\begin{equation}
    \ATTN(\TRFH)
    =
    \SOFTMAX\!\left(\frac{\TRFQ \TRFK^\top}{\sqrt{d_k}}\right)\TRFV. 
\end{equation}

The attention matrix assigns a data-dependent weight to each pair of input elements, allowing each representation to aggregate information from all others. Multi-head attention repeats this computation with several learned projections and concatenates the resulting outputs, allowing the model to capture different types of interactions in parallel.

A transformer layer typically combines multi-head self-attention with a position-wise feed-forward network, residual connections, and normalization. For stacked transformer layers, let $\TRFH^{(l)}$ denote the sequence of embeddings at layer $l$. A transformer layer computes
\begin{equation}
    \tilde{\TRFH}^{(l)}
    =
    \NORM\!\left(
        \TRFH^{(l)} + \MHA(\TRFH^{(l)})
    \right),
\end{equation}
\begin{equation}
    \TRFH^{(l+1)}
    =
    \NORM\!\left(
        \tilde{\TRFH}^{(l)} + \FFN(\tilde{\TRFH}^{(l)})
    \right),
\end{equation}
where $\MHA$ denotes multi-head attention and $\FFN$ denotes the position-wise feed-forward network, implemented as an MLP applied independently to each element of the sequence. Figure~\ref{fig:background_transformer_layer} summarizes the structure of a generic transformer layer.

Although transformers were first developed for sequences, the same mechanism can be applied to sets or graph-derived collections of embeddings when suitable structural or positional information is provided. From this viewpoint, self-attention acts as a learned aggregation operator over a dense dependency structure.

Graph attention and graph-transformer models adapt this idea to graph-structured data by combining attention with graph-specific encodings or local message passing~\cite{velickovic_graph_2017,rampasek_recipe_2022}. Later in the thesis, this attention-based view will be used as one possible mechanism for remapping graph embeddings to other structured representations.

\input{resources/tikz/chapter_1/ml/transformer_feature_mapping}

\paragraph{Graph features for combinatorial instances}

When applying a graph neural or attention-based model to a combinatorial optimization instance, the graph structure specifies which nodes exchange messages, but the model also requires numerical features on nodes and edges. In the variable-constraint graph of a BILP, variable-node features may include objective coefficients, LP-relaxation values, reduced costs, or basis-state information. Constraint-node features may include right-hand-side values, slack values, active-constraint indicators, or dual values. Edge features typically encode the coefficient of a variable in a constraint, possibly after normalization.

In addition to optimization-derived quantities, one may use purely structural graph features. Degree features describe local connectivity, while positional encodings such as Laplacian eigenvectors or random-walk features provide information about a node's position in the graph beyond its immediate neighborhood~\cite{maskey_generalized_2022,eliasof_graph_2023}. 

Random node features can also be added to break symmetries or increase the expressive power of message-passing GNNs~\cite{sato_random_2021,abboud_surprising_2021}. These features are used as inputs to neural encoders for variable scoring and MRF parameter prediction in Chapter~\ref{chapter:3_differentiable_mrf_inference_for_BILPs}, denoising updates in Chapter~\ref{chapter:4_denoising_and_mrf_inference}, and feature-to-factor mappings in Chapter~\ref{chapter:5_feature-to-factor_mappings_for_approximate_mrf_inference}.

\subsection{Summary: Learning structured predictors}

This section introduced the machine-learning concepts used in the remainder of the thesis. Supervised learning provides the basic framework for training direct variable-level predictors and learned MRF parameterizations from reference solutions. Binary cross-entropy is used for direct binary prediction, while Chapter~\ref{chapter:3_differentiable_mrf_inference_for_BILPs} uses inference-based surrogate losses to train structured MRF predictors.

We also introduced denoising and flow-matching objectives as generative approaches to structured solution construction. These methods view solution generation as an iterative refinement process, transforming noisy, relaxed, or partially specified assignments into candidate solutions. This perspective is important for Chapter~\ref{chapter:4_denoising_and_mrf_inference}, where denoising and flow-based models are combined with MRF-structured predictions.

Finally, GNNs and transformers provide the neural representation and mapping mechanisms used to process variable-constraint graphs and construct learned predictors. GNNs encode the local structure of the original ILP representation, while attention-based modules provide flexible ways to aggregate and remap embeddings, a role that becomes important in the feature-to-factor mappings of Chapter~\ref{chapter:5_feature-to-factor_mappings_for_approximate_mrf_inference}. Section~\ref{sec:mrf_intro} introduces Markov Random Fields and approximate inference, which supply the structured probabilistic layer used to model dependencies between BILP decision variables.