\chapter{Differentiable MRF Inference for Binary Integer Linear Programs}
\label{chapter:3_differentiable_mrf_inference_for_BILPs}
\chaptermark{Differentiable MRFs}

\section{Introduction}
\label{sec:chapter_3_introduction}

Binary integer linear programs (BILPs) provide a common mathematical formulation for combinatorial optimization tasks in which the decision variables are binary and feasibility is defined by linear constraints. This chapter focuses on BILP solution prediction: given a BILP instance, the goal is not to replace an exact solver, but to produce structured prediction signals over the binary variables that can guide a downstream decoder or solver. The broader difficulty of large-scale combinatorial optimization has motivated learning-based methods that predict high-quality assignments, construct solutions sequentially, or guide exact solvers toward promising regions of the search space~\cite{dai_learning_2017,gasse_exact_2019,nair_solving_2020,sun_difusco_2023}. Here, we restrict attention to the BILP setting, where the variable--constraint structure of the instance provides a natural graph representation for neural prediction.

For linear and integer linear programs more generally, a common strategy is to represent the instance as a variable--constraint bipartite graph and apply a GNN to produce variable-level scores. In the BILP setting considered here, these scores are usually interpreted as independent probabilities that each binary variable should be set to one in an optimal solution. This factorized prediction model is simple and scalable, and it forms the basis of several successful neural approaches. However, it has an important limitation: at decoding time, dependencies between variables are only represented implicitly through the GNN embeddings. The probabilistic model itself does not explicitly encode the constraints that couple variables together. As a result, independent variable predictions can lead to infeasible or low-quality assignments, requiring additional rounding, greedy repair, or solver-based post-processing.

This chapter addresses this limitation by replacing factorized variable predictions with a structured probabilistic model. We use MRFs to represent dependencies between decision variables through constraint factors. In this view, the decision variables of the BILP become random variables in an MRF, while constraints define factors over subsets of variables. Approximate inference algorithms, such as LBP and NMF, can then be used to compute marginals that reflect not only the score of each variable but also its compatibility with neighboring variables and constraints. These marginals are not themselves complete feasible solutions; they are structured signals that guide the decoding and feasibility-propagation procedures described in Section~\ref{subsubsec:chapter_3_egg_gd}.

The approach developed in this chapter builds on the neural prediction framework of Nair et al.~\cite{nair_solving_2020}, but changes the probabilistic model used after the GNN encoder. Instead of predicting a fully factorized distribution over variables, the model uses an exponential-family MRF over the BILP decision variables. The non-learned canonical Markov random field (CMRF), defined in Section~\ref{subsec:canonical-feasibility-cost-param}, serves as a structured baseline: its local log-potentials are fixed directly by the BILP constraints and objective. The learned MRF (LMRF) studied in this chapter keeps the same factor graph and the same over-complete sufficient-statistic representation, but replaces the fixed canonical parameter vector $\CANEPARAM(\PROB)$ with predicted parameters $\PREEPARAM(\PROB;\NNPARAM)$. In this sense, the learned MRF models can be viewed as relaxations of the canonical feasibility-cost construction: the structure of the MRF is retained, while the log-potential values are learned from data.

A key advantage of this formulation is that fixed-iteration inference can be implemented as a differentiable computation. The output marginals can
therefore be used both for training and for decoding. During training, approximate marginals define a surrogate objective that encourages the model
to assign high probability to observed high-quality solutions. During decoding, the same inference procedure provides structured variable marginals that guide solution construction. We use these marginals both to rank variables and, more importantly, in a guided diving strategy that progressively fixes variables and recomputes the remaining marginals under the accumulated evidence.

We evaluate this framework on two synthetic BILP families: MIS and SC. These problems provide complementary test cases. MIS has small pairwise constraints and a maximization objective, while SC has higher-order covering constraints and a minimization objective. Across these settings, the experiments compare factorized GNN predictors, non-learned canonical MRF baselines, and learned MRF predictors trained through approximate inference. The results indicate that learned MRF inference can provide a stronger decoding signal than factorized prediction on these benchmarks, especially when inference is coupled with adaptive decoding through guided decimation.

The contributions of this chapter are as follows:
\begin{enumerate}
    \item We formulate BILP prediction signals as approximate inference outputs in an exponential-family MRF defined over the variables and constraints of a BILP.

    \item We extend the background CMRF construction to learned over-complete MRF parameterizations, where a GNN predicts factor parameters and fixed-iteration LBP or NMF produces approximate marginals.

    \item We study diving-based decoders that convert inferred marginals into feasible solutions, including Entropy-Guided Greedy (EGG) and Guided Decimation (GD). GD differs from EGG by conditioning the MRF on partial assignments and recomputing marginals during the dive.

    \item We empirically evaluate the resulting CMRF and LMRF models on synthetic MIS and SC BILP instances, analyzing solution quality, inference budget, factor-potential behavior, and the computational cost of structured inference.
\end{enumerate}

\subsection{Positioning with respect to related work}

The broader landscape of learning-based optimization methods and graphical-model approaches was reviewed in Chapter~\ref{chap:chapter_2_related_work}.
Here, we focus only on the works most directly connected to the MRF-based BILP predictor developed in this chapter.

The closest starting point is the class of neural methods that represent an optimization instance by its variable--constraint graph and predict variable-level scores. In particular, the framework of Nair et al.~\cite{nair_solving_2020} uses a neural predictor to guide the search for high-quality solutions. Similar graph-based learning ideas have also been used for construction heuristics and solver guidance \cite{dai_learning_2017,gasse_exact_2019}. These approaches demonstrate the usefulness of learned variable scores, but the prediction model used for decoding is typically factorized across variables. The method in this chapter keeps the same graph-based neural input representation, but replaces the factorized output distribution with an MRF over binary assignments.

A second line of work studies message passing as an optimization procedure. Park and Shin~\cite{park_max-product_2017} show that max-product BP can solve certain combinatorial optimization problems when the corresponding graphical model and LP relaxation satisfy specific assumptions. This result motivates the canonical feasibility-cost parameterization reviewed in Section~\ref{subsec:canonical-feasibility-cost-param}, where constraint factors encode local feasibility and unary factors encode the linear objective. In contrast to the exact-recovery setting analyzed by Park and Shin, the BILP instances considered here are treated with approximate finite-iteration inference.

Our approach is also related to works that unroll or learn BP-style inference procedures \cite{kuck_belief_2020,deng_deep_2022,wiseman_amortized_2019}. Those methods typically aim to improve the inference algorithm itself, for example by learning message updates, damping, or schedules. Here, LBP and NMF are used differently: they are fixed approximate inference layers applied to an MRF whose parameters are either given by the canonical BILP construction or predicted by a neural model. The resulting approximate marginals are used for training and decoding. They should therefore be understood as surrogate inference outputs, not as exact likelihoods or exact marginals.

Finally, the framework studied in this chapter differs from generative, sampling-based, and solver-in-the-loop approaches such as diffusion or perturb-and-MAP methods \cite{sun_difusco_2023,berthet_learning_2020}. Instead of relying on Monte Carlo estimates, denoising trajectories, or repeated solver calls during training, the models studied here use deterministic fixed-iteration LBP or NMF to produce structured marginals. This makes the chapter's contribution specifically about replacing factorized BILP prediction with differentiable MRF-based prediction and decoding.



\section{Learning MRF parameters for BILPs}
\label{sec:chapter_3_mrf_parameter_learning}

In this section we describe the framework used in this chapter, combining GNNs with MRF inference. The central idea is to move beyond fully factorized predictors, which treat variable assignments as independent, and instead define a structured distribution over binary assignments. This distribution is modeled as an exponential-family MRF whose parameters are either fixed from the BILP formulation or predicted by a GNN encoder operating on the variable--constraint graph. Inference in this MRF---via belief propagation or mean-field approximations---yields marginal distributions that capture dependencies between variables and can be used to guide diving-based solution construction. We consider two decoders: Entropy-Guided Greedy (EGG), which computes scores once and then performs a static dive, and Guided Decimation (GD), which periodically conditions the MRF on the current partial assignment and reinfers the remaining marginals.

We first define the generic BILP representation used by the model, then recall the two benchmark families used in the experiments.

\subsection{Problem class and graph representation}
In this thesis we focus on optimization problems that can be formulated as a BILP of the form
\begin{equation}
    (\PROB) \Leftrightarrow 
    \left\{ 
    \begin{array}{ll}
         \max        & \CVEC^\top \XVEC \\
         \text{s.t.} & \AMAT \XVEC \leq \BVEC \\
                     & \XVEC \in \{0,1\}^{\NVAR},
    \end{array}
    \right.
\end{equation}
with objective coefficients $\CVEC \in \R^\NVAR$ and constraints defined by $(\AMAT,\BVEC) \in \R^{\NCONSTR \times \NVAR} \times \R^{\NCONSTR}$. We denote this family of BILP instances by $\ILPFAM$. Minimization problems are handled by the equivalent maximization form obtained by replacing the objective coefficients by their negatives. This convention allows us to use a single notation throughout the MRF construction; when reporting experimental gaps, however, we use the original objective sense of each problem.


For a given instance $(\PROB)$, a solution $\XVEC$ can be viewed as a realization of a binary random vector $X=(X_i)_{i=1}^{\NVAR}$. Conceptually, one may associate the instance with an unknown target distribution over high-quality or optimal assignments,
\begin{equation*}
p^*(\XVEC \mid \PROB).
\end{equation*}
In practice, this distribution is not observed. The training data provide one reference solution, or a small set of reference solutions, for each instance. The learned model is therefore trained to reproduce local information from the available reference solutions, rather than to estimate the full target distribution exactly.

A common neural approach is to represent $(\PROB)$ by its variable--constraint bipartite graph and use a GNN to produce one score $\SCORE_i$ for each decision variable \cite{dai_learning_2017,gasse_exact_2019,nair_solving_2020,sun_difusco_2023}. See Section~\ref{sec:vcg_definition} for details on this construction. In the binary setting, these scores are typically interpreted as Bernoulli probabilities
\begin{equation*}
    \SCORE_i \approx \PDIST(\XRAV_i = 1 \mid \PROB).
\end{equation*}
This gives the fully factorized predictive distribution
\begin{equation}
    \label{equ:fully_factorized_distribution}
    \PDIST_{\mathrm{fac}}(\XVEC \mid \PROB)
    =
    \prod_{i=1}^{\NVAR}
    \PDIST_i(\XRAV_i = \XVEC_i \mid \PROB).
\end{equation}
The GNN encoder may incorporate information from constraints through message passing on the input graph, but the output distribution $\PDIST_{\mathrm{fac}}$ treats the binary variables as conditionally independent given the instance. Dependencies between variables are therefore represented only implicitly in the embeddings, not explicitly in the probabilistic model used for decoding.

The central change in this chapter is to replace the factorized Bernoulli model in~\eqref{equ:fully_factorized_distribution} with an MRF over the same binary decision variables. Let
\begin{equation*}
    \VCGFULL
\end{equation*}
denote the variable--constraint bipartite graph used as the input representation of the BILP instance, where $\VCGV$ indexes decision variables
and $\VCGC$ indexes constraints. Separately, let 
\begin{equation*}
    \FGFULL
\end{equation*}
denote the factor graph on which the MRF distribution is defined, where $\FGF$ is the set of MRF factors. In this chapter, each BILP constraint gives rise to one factor, so the two graphs have the same bipartite structure and we may identify $\FGF$ with $\VCGC$. We keep the notation distinct because $\VCGFULL$ refers to the graph used by the neural encoder, whereas $\FGFULL$ refers to the probabilistic graphical model used for inference.

The structured model takes the form
\begin{equation}
    \label{equ:mrf_distribution_intro}
    \PDIST_{\EPARAM}(\XVEC \mid \PROB)
    =
    \frac{1}{Z_{\EPARAM}(\PROB)}
    \prod_{\FACTOR \in \FGF}
    \POTENTIAL_{\FACTOR}
    \left(\XVEC_{\FACTOR};\EPARAM_{\FACTOR}(\PROB)\right),
\end{equation}
where $\XVEC_{\FACTOR}$ denotes the variables in the scope of factor $\FACTOR$, $\POTENTIAL_{\FACTOR}$ encodes local compatibility, and $Z_{\EPARAM}(\PROB)$ is the partition function. Unlike the factorized model, this distribution explicitly couples variables that participate in the same constraint factor. Approximate inference in this MRF, using LBP or NMF, produces marginals that can then be used for training and decoding.

Figure~\ref{fig:model_design_nair} illustrates the factorized baseline of Nair et al.~\cite{nair_solving_2020}. The following sections introduce the MRF-based alternative: first the exponential-family formulation used for BILP prediction, then the learned parameter prediction architecture. The CMRF baseline uses the fixed canonical parameters defined in Section~\ref{subsec:canonical-feasibility-cost-param}, whereas the LMRF variants replace them with predicted parameters.

\input{resources/tikz/chapter_3/model_design_nair}

\paragraph{Benchmark problem families.}
We evaluate the MRF predictor on the MIS and SC BILP families introduced in Section~\ref{subsec:background_representative_problems}. We reuse the notation and formulations from that background section: MIS is considered in the edge-based formulation of~\eqref{equ:mis_formulation_edges}, while SC is considered in the standard covering formulation of~\eqref{equ:sc_formulation}. For MIS, the choice of the edge-based formulation is important for the graphical model studied here. Each constraint $\XVEC_i+\XVEC_j\leq 1$ involves exactly the two endpoint variables of an edge, so the induced MRF contains only arity-$2$ constraint factors. This pairwise factor structure keeps the over-complete factor potentials small and makes LBP and NMF inference directly comparable to the binary-factor construction used throughout this chapter. By contrast, using a clique formulation for MIS would replace edge constraints by larger clique factors and would change the factor arities of the resulting MRF.

For SC, each item-covering constraint gives rise to one MRF factor whose arity is the number of candidate sets covering that item. In the synthetic SC instances used below this arity is controlled by the instance-generation parameter $k$, so the same MRF construction can be evaluated on covering constraints while keeping factor sizes manageable. Thus, the two benchmark families test complementary structures: pairwise exclusion factors for edge-based MIS and controlled higher-order covering factors for SC.

The features provided to the model are extracted by solving the continuous relaxation of the corresponding BILP. For the edge-based MIS formulation, the relaxation admits the uninformative solution $\XVEC_i=1/2$, so the learned model must rely primarily on graph topology rather than on informative LP feature values.

\subsection{Exponential-family MRF parameterization}
For a given problem $(\PROB)$, we consider an MRF defined on a factor graph $\FGFULL$, whose factors are indexed by $\FGF$. Each factor $\FACTOR \in \FGF$ has scope $N(\FACTOR)$ and local assignment $\XVEC_{\FACTOR}$. We restrict attention to exponential-family factors of the form
\begin{equation}
\POTENTIAL_{\FACTOR}(\XVEC_{\FACTOR})
=
\exp\!\left(
    \left\langle
    \EPARAM_{\FACTOR},
    \SSTAT_{\FACTOR}(\XVEC_{\FACTOR})
    \right\rangle
\right),
\end{equation}
so that the overall distribution is
\begin{equation}
    \PDIST_{\EPARAM}(\XVEC \mid \PROB)
    \propto
    \exp \!\left(
        \sum_{\FACTOR \in \FGF}
        \left\langle
        \EPARAM_{\FACTOR},
        \SSTAT_{\FACTOR}(\XVEC_{\FACTOR})
        \right\rangle
    \right).
\end{equation}
Here $\EPARAM = (\EPARAM_{\FACTOR})_{\FACTOR \in \FGF}$ is the collection of factor parameters and $\SSTAT(\XVEC) = (\SSTAT_{\FACTOR}(\XVEC_{\FACTOR}))_{\FACTOR \in \FGF}$ is the collection of sufficient statistics. For each factor, $\SSTAT_{\FACTOR}(\XVEC_{\FACTOR}), \EPARAM_{\FACTOR} \in \mathbb{R}^{\FDIM_{\FACTOR}}$, and the total parameter dimension is
\begin{equation*}
    \SSDIM = \sum_{\FACTOR \in \FGF} \FDIM_{\FACTOR}.
\end{equation*}
In this parameterization, learning an MRF amounts to choosing the factor parameters $\EPARAM = (\EPARAM_{\FACTOR})_{\FACTOR \in \FGF}$ so that the resulting distribution assigns high probability to assignments that resemble the training data.

Throughout this chapter, we work in log-potential form. Equivalently, the parameter score can be interpreted as a negative local energy:
\begin{equation*}
    \langle \EPARAM_{\FACTOR},
    \SSTAT_{\FACTOR}(\XVEC_{\FACTOR}) \rangle
    =
    -E_{\FACTOR}(\XVEC_{\FACTOR}).
\end{equation*}
Thus larger values of the parameter score correspond to lower energy and higher probability. MAP inference therefore maximizes total negative energy, or equivalently minimizes total energy.

\paragraph{Canonical over-complete sufficient statistics.} We use the \emph{canonical over-complete representation}~\cite{wainwright_graphical_2007} (see Section~\ref{sec:mrf_intro}). For each factor $\FACTOR \in \FGF$, the over-complete indicator representation contains one sufficient statistic for each local assignment $\XVEC_{\FACTOR}$. Since the variables are binary, this gives
\begin{equation*}
    \FDIM_{\FACTOR} = 2^{|\FACTOR|}.
\end{equation*}
The total number of parameters of the MRF corresponding to a problem $(\PROB)$ is denoted $\SSDIM$. The associated factor parameter vector
$\EPARAM_{\FACTOR} \in \R^{\FDIM_{\FACTOR}}$ has one entry per local assignment. We refer to these as \emph{over-complete factor parameters} as shorthand: the over-completeness belongs to the sufficient-statistic representation, while the parameters are indexed by that representation.



\subsection{Neural parameter predictor}
The LMRF uses a neural network to predict the over-complete factor parameters of the MRF associated with a problem instance $(\PROB)$. The neural network encodes the variable--constraint graph $\VCGFULL$, while the predicted parameters are assigned to factors in the MRF factor graph $\FGFULL$. Our predictor combines two components:
\begin{enumerate}
    \item \textbf{Graph encoder.} A multi-layer GNN operating on the variable–constraint bipartite graph computes embeddings $\EMB_u \in \R^{\EMBDIM}$ for each vertex $u \in \VCGV \cup \VCGC$. This serves as feature extraction from the structured problem representation.
    \item \textbf{Parameter predictor.} Each MRF factor is assigned an input embedding derived from the corresponding vertex of the variable--constraint graph. Since the MRF contains one unary factor for each decision variable and one non-unary factor for each constraint, we use the variable-node embedding as the input embedding for the corresponding unary factor, and the constraint-node embedding as the input embedding for the corresponding constraint factor. Denoting this factor input embedding by $\FEMB$, a differentiable function $\FPARAMPRED_{\FACTOR}$ produces \(\PREEPARAM_{\FACTOR}\), the learned factor-parameter table for factor \(\FACTOR\):
\begin{equation*}
\PREEPARAM_{\FACTOR} = \FPARAMPRED_{\FACTOR}(\FEMB).
\end{equation*}
\end{enumerate}
The LMRF models use the predicted parameter vector $\PREEPARAM(\PROB;\NNPARAM)$, where $\NNPARAM$ denotes the trainable weights of the encoder and parameter heads. Figure~\ref{fig:model_design_efcanonicalparam} illustrates this architecture.

\paragraph{Graph encoder.} A problem $(\PROB)$ is represented as its variable--constraint bipartite graph $\VCGFULL$ (where $\VCGV$ indexes the set of decision variables and $\VCGC$ the set of constraints) and by a set of features $\FEATURES = \{ \FEATURES_u \}_{u \in \VCGV \cup \VCGC}$. The graph encoder is a parameterized differentiable function that converts the structured representation of the problem into a set of vector embeddings $\EMB = \{ \EMB_u \}_{u \in \VCGV \cup \VCGC }$ that serve as inputs to the following components of the predictor. We denote the set of trainable parameters of the graph encoder $\GNNEMB$ by $\GNNEMBPARAM$.
\begin{equation}
    \GNNEMB (\VCG, \FEATURES; \GNNEMBPARAM) = \EMB.
\end{equation}

In practice $\GNNEMB$ is defined as a sequence of $\NUMGNNLAYERS$ graph convolution layers (see experimental results for details on the GNN implementation). 

\paragraph{Parameter prediction head.} The parameter prediction head transforms the vertex embeddings produced by the graph encoder into factor-specific input embeddings and then into predicted over-complete factor parameters $\PREEPARAM(\EMB)$. We denote this function as $\FPARAMPRED$ and its set of trainable parameters $\FPARAMPREDPARAM$. For each factor arity $k \in \{1,\ldots,\max_{\FACTOR\in\FGF}|\FACTOR|\}$, we define a predictor
$\FPARAMPRED_k : \R^{\EMBDIM} \rightarrow \R^{2^k}$. Thus, for a factor $\FACTOR$ with $k=|\FACTOR|$ adjacent binary variables, the head outputs one log-potential parameter for each of the $2^k$ local assignments.

The final parameter vector is obtained as 
\begin{equation}
    \PREEPARAM = \{\FPARAMPRED_{|\FACTOR|} (\FEMB; \FPARAMPREDPARAM)\}_{\FACTOR\in \FGF}.
\end{equation}

Unary and non-unary factors are handled by the same over-complete representation, but correspond to different factor arities. For a unary factor associated with variable $i$, the predictor outputs two parameters, one for $\XVEC_i=0$ and one for $\XVEC_i=1$. For a non-unary constraint factor $\FACTOR$, the predictor outputs one parameter for every joint assignment of the variables in $N(\FACTOR)$. These non-unary tables allow the LMRF to learn interactions between variables that participate in the same constraint.

Unlike the CMRF, the LMRF does not impose a fixed $-M$ penalty or a hard mask on infeasible local assignments. Feasible and infeasible local configurations both receive finite learned parameters. Therefore, global feasibility is not guaranteed by the learned MRF distribution itself; it must be handled by the decoding or repair procedure.

\subsection{Differentiable inference}
\label{subsec:chapter_3_differentiable_inference}

Once parameters are predicted, inference in the MRF provides structured predictions. Depending on the task, we may:
\begin{itemize}
    \item compute approximate marginals via LBP or NMF,
    \item perform conditional marginal inference when partial assignments are known, as in guided decimation,
    \item or, in the canonical MRF interpretation, view exact MAP inference as the idealized optimization problem encoded by the model.
\end{itemize}
Inference thus bridges learning and optimization by allowing predictions to take variable dependencies into account.

It is worth noting the conceptual similarities with recent approaches that integrate combinatorial optimization problems as layers within machine learning models \citep{berthet_learning_2020,dalle_learning_2022}. In these works, the optimization layer is treated as a black box: logits or continuous parameters are passed to a state-of-the-art solver that computes exact (but non-differentiable) solutions, while gradient information is approximated by sampling or perturbing the inputs to estimate the sensitivity of the outputs. In contrast, this chapter uses approximate marginal inference as the inference mechanism. We instantiate this mechanism with LBP and NMF, both of which produce differentiable approximate marginals through fixed-iteration message-passing or variational updates. These marginals define a smooth energy-based relaxation over assignments, which can be used for decoding while remaining amenable to end-to-end training and efficient parallelization. In the canonical parameterization, exact MAP inference in the resulting MRF would correspond to solving the encoded combinatorial optimization problem. In this chapter, however, we use approximate marginal inference as a differentiable surrogate that guides decoding rather than directly solving the MAP problem exactly.

At this stage, the model has computed the parameters associated with instance $\PROB$, denoted $\PREEPARAM(\PROB;\NNPARAM)$. Training and decoding then require approximate inference in the resulting MRF. During training, inference returns factor or variable marginals used in the supervised surrogate loss. During decoding, inference returns variable marginals used by the diving procedures. Because fixed-iteration LBP and NMF consist of differentiable operations, gradients can be propagated through the inference layer during end-to-end training.

This chapter uses two inference methods: LBP and NMF; see Sections~\ref{subsec:background_lbp} and~\ref{subsec:background_mean_field} for details. The advantage of these inference methods is that they incorporate dependencies between variables while still producing differentiable marginal estimates. They also allow the MRF to be conditioned on partial evidence during decoding, which is the mechanism used by guided decimation. This conditioning mechanism is used by the guided-decimation procedure described in Section~\ref{subsubsec:chapter_3_egg_gd}.

The messages, variational updates, and marginals computed by fixed-iteration LBP and NMF are differentiable operations and do not introduce additional trainable parameters. Because both LBP and NMF are iterative approximate inference procedures, we use a fixed maximum number of iterations and optionally apply update damping for stability \cite{heskes_uniqueness_2004,wainwright_graphical_2007,meltzer_convergent_2012}. In LBP, damping interpolates between old and new messages. In NMF, we apply the same relaxation idea to variable-marginal updates. 

\paragraph{Conditioning on partial assignments.} During decoding, we sometimes condition the MRF on a partial assignment $\PARTSOL$. This is implemented by modifying the local factor tables: for each factor, local assignments that disagree with the fixed variables in $\PARTSOL$ are masked by assigning them log-potential $-\infty$, or a large negative value in finite-precision implementations. Approximate inference on the resulting conditioned MRF estimates marginals of the remaining variables under the accumulated evidence.

\subsection{Solution decoding}
\label{subsubsec:chapter_3_egg_gd}
Having predicted the parameters $\PREEPARAM(\PROB)$ and run approximate inference, we must decode the resulting scores or marginals into a feasible solution of the original problem $(\PROB)$. This section describes the decoding procedures used in our experiments: Entropy-Guided Greedy and Guided Decimation. EGG constructs a solution from a single set of scores or marginals, while GD alternates between variable fixing and conditional reinference.

\subsubsection{Diving-based decoding}

We decode solutions using diving procedures. A dive maintains a partial assignment $\PARTSOL$ and repeatedly fixes one candidate variable to a selected value. After each fixing step, the remaining problem is simplified by propagating the consequences of the decision: variables that can no longer take one of their values without violating feasibility are fixed or removed from the candidate set. This propagation keeps the partial assignment feasible at every step.

In this chapter we use two diving decoders. EGG performs a single inference pass at the beginning and uses the resulting marginals or scores throughout the entire dive. GD uses the same diving procedure, but periodically conditions the MRF on the current partial assignment and reruns inference to update the scores used for subsequent decisions.

\paragraph{Feasibility propagation for MIS and SC.} The propagation step is problem-specific. For MIS, fixing a vertex to one forces all adjacent vertices to zero, since they can no longer be included in the independent set. Vertices fixed to zero are removed from the candidate set. For SC, fixing a set to one covers its elements and reduces the residual covering problem; sets that no longer cover any uncovered element can be discarded, while elements that are covered by only one remaining set force that set to be selected. These deterministic updates reduce the problem after each dive step and maintain feasibility of the partial assignment.

\input{resources/tikz/chapter_3/diving_steps}

\subsubsection{Entropy-Guided Greedy and Guided Decimation}

Both EGG and GD use the same diving and feasibility-propagation mechanism. They differ only in how the scores used for selecting the next fixing decision are obtained. EGG computes marginals or scores once on the original instance and keeps them fixed during the whole dive. GD uses the conditioning operation described above after each fixing step, or after a fixed number of steps, to update the marginals used for subsequent decisions.

Whereas predicting scores with the GNN does not provide any way of integrating partial assignments into the marginal calculations, MRFs do. When building a solution, we can incorporate the partial solution as evidence into inference and update the marginals used in subsequent construction steps. Details of EGG and GD iterations are given in Algorithm~\ref{alg:bp_guided_decimation}.

\begin{figure}[htbp]
  \centering
  \begin{minipage}{.7\linewidth}
    \begin{algorithm}[H]
        \LinesNumbered
        \caption{Diving decoder with optional guided decimation}
        \KwData{$\PROB$, MRF parameters $\EPARAM$, inference method $\mathcal{I}$, update interval $k$}
        \KwResult{Feasible solution $\XVEC$}
        $\PARTSOL \gets \emptyset$ \tcp*{partial assignment}
        Initialize candidate set $C$\;
        Run inference $\mathcal{I}$ on the original MRF to obtain marginals $\MARGINAL$\;
        \While{solution incomplete}{
            %Select a candidate fixing $(i, \XVEC_i)$ using the current scores $\MARGINAL$\;
            Select an unfixed variable $i \in \argmin_{j \in C} H(\MARGINAL_j)$\;
            Set $\XVEC_i \gets \argmax_{b \in \{0,1\}} \MARGINAL_i(b)$\;
            Add $(i, \XVEC_i)$ to $\PARTSOL$\;
            Propagate feasibility consequences and update $C$\;
            \If{using GD and an update is triggered}{
                Condition the MRF on $\PARTSOL$ by masking incompatible local assignments\;
                Rerun inference $\mathcal{I}$ to update marginals $\MARGINAL$\;
            }
        }
        Return $\XVEC$ induced by $\PARTSOL$\;
    \end{algorithm}
    \label{alg:bp_guided_decimation}
    \end{minipage}
\end{figure}

The decoder uses an entropy-guided greedy selection rule. At each step, it selects the unfixed variable whose current approximate marginal has minimum entropy, i.e., the variable for which inference is most certain. For a binary variable with marginal $\MARGINAL_i$, this entropy is
\begin{equation}
    H(\MARGINAL_i)
    =
    - \MARGINAL_i(0)\log \MARGINAL_i(0)
    - \MARGINAL_i(1)\log \MARGINAL_i(1).
\end{equation}
The selected variable is then assigned to its most likely value,
\begin{equation}
    x_i \gets \argmax_{b \in \{0,1\}} \MARGINAL_i(b).
\end{equation}
EGG applies this rule using the initial scores throughout the dive. GD applies the same rule, but after a fixed number of fixing steps it conditions the MRF on the current partial assignment and reruns inference on the residual problem. Cheaper update schedules are evaluated experimentally in Section~\ref{sec:chapter_3_experiments}.

GD addresses a limitation of EGG. Because EGG uses the initial scores throughout the dive, it may continue to favor variables whose relevance has changed after earlier fixings and feasibility propagation. GD mitigates this by conditioning on the current partial assignment and recomputing the remaining marginals.

\subsection{Training objective}
\label{sec:chapter_3_training_objective}

The learned MRF models are trained from supervised solution labels. For each training instance, the neural network first predicts MRF parameters from the variable--constraint graph. Approximate inference is then run on the resulting MRF, producing either factor marginals with LBP or variable marginals with NMF. The training loss compares these approximate marginals to a reference solution. Importantly, this loss is not the exact negative log-likelihood of the full MRF
distribution; it is a tractable supervised marginal-likelihood surrogate defined on the outputs of the chosen inference procedure.

\paragraph{Model parameters} 
Throughout this section, $\NNPARAM$ denotes the trainable neural parameters of the learned MRF model, including the GNN encoder parameters and the factor-parameter prediction heads,
\begin{equation*}
    \NNPARAM = (\GNNEMBPARAM, \FPARAMPREDPARAM).
\end{equation*}
Given an instance $\PROB$, these parameters define predicted MRF parameters $\PREEPARAM(\PROB;\NNPARAM)$,which are then used by LBP or NMF to compute approximate marginals. The symbol $\CANEPARAM(\PROB)$ is reserved for the fixed canonical feasibility-cost parameters used by the non-learned CMRF baselines.

We train the model in a supervised learning setting. Given a dataset
\begin{equation*}
    \mathcal{D} = \{(\PROB_i,\YVEC_i)\}_{i=1}^{N},
\end{equation*}
where $\PROB_i$ is a problem instance and $\YVEC_i$ is a reference solution, we optimize the trainable neural parameters
\begin{equation}
    \NNPARAM^*
    =
    \argmin_{\NNPARAM}
    \frac{1}{N}
    \sum_{i=1}^{N}
    \LOSSF_i(\NNPARAM).
\end{equation}

For LBP-based models, inference returns approximate factor marginals. We use the factor-wise marginal negative log-likelihood surrogate
\begin{equation}
\LOSSF_i^{\mathrm{LBP}}(\NNPARAM)
=
-\frac{1}{|\FGF_i|}
\sum_{\FACTOR \in \FGF_i}
\log
\MARGINAL_{\FACTOR}^{(i)}
\left(
    \YVEC_{i,\FACTOR}
\right),
\end{equation}
where $\MARGINAL_{\FACTOR}^{(i)}$ is the approximate marginal of factor $\FACTOR$ obtained after running LBP with parameters $\PREEPARAM(\PROB_i;\NNPARAM)$.

For NMF-based models, the approximating distribution is fully factorized and inference returns variable marginals. We therefore use the variable-wise surrogate
\begin{equation}
\LOSSF_i^{\mathrm{NMF}}(\NNPARAM)
=
-\frac{1}{|\VCGV_i|}
\sum_{v \in \VCGV_i}
\log
\MARGINAL_{v}^{(i)}
\left(
    \YVEC_{i,v}
\right),
\end{equation}
where $\MARGINAL_{v}^{(i)}$ is the approximate marginal of variable $v$ obtained after running NMF with parameters $\PREEPARAM(\PROB_i;\NNPARAM)$.

\paragraph{Why this loss?} 
The purpose of this objective is to train the neural parameterization through the marginals produced by the chosen inference procedure, while avoiding the intractable normalization required by the full MRF likelihood. Instead of assigning probability to an entire solution $\YVEC$ in the exponentially large assignment space, the loss compares local approximate marginals to the corresponding local parts of the reference solution. It should therefore be understood as a tractable supervised marginal-likelihood surrogate, not as exact maximum likelihood training of the MRF.

\paragraph{LBP versus NMF} 
The form of the local surrogate depends on the inference method. With LBP, inference returns approximate factor marginals, so each factor $\FACTOR$ defines a multi-class prediction problem over the $2^{|\FACTOR|}$ local configurations. With NMF, the approximating distribution is fully factorized and only variable marginals are available, so the loss reduces to variable-level binary classification.

\paragraph{Relation to maximum likelihood}
The natural training target is to minimize the global negative log-likelihood (NLL) of the MRF:
\begin{align}
\mathcal{L}_{\mathrm{NLL}}(\NNPARAM) &= -\log p_{\PREEPARAM(\PROB;\NNPARAM)}(\YVEC), \\
                                   &= \log Z(\PREEPARAM(\PROB;\NNPARAM)) - \langle \PREEPARAM(\PROB;\NNPARAM), \SSTAT(\YVEC) \rangle.
\end{align}
This objective is not used in this chapter, because evaluating it requires the partition function $Z(\PREEPARAM(\PROB;\NNPARAM))$. For the loopy factor graphs considered here, this normalization is intractable. The losses above therefore serve as supervised marginal-likelihood surrogates: they train the model to assign high approximate marginal probability to the local configurations observed in the reference solution, rather than maximizing the exact likelihood of the full assignment. A principled alternative is to approximate the likelihood using variational free energy methods such as the Bethe free energy, which provide tractable approximations consistent with loopy BP. Exploring such objectives is a promising avenue for future work.

Gradients are computed by backpropagating through the fixed unrolled inference computation. For LBP, this computation consists of a fixed number of differentiable message updates followed by marginal normalization. For NMF, it consists of a fixed number of differentiable mean-field marginal updates. Since the number of iterations is fixed, the inference procedure defines a deterministic differentiable map from $\PREEPARAM(\PROB_i;\NNPARAM)$ to the marginals used in the loss.

\paragraph{Limitations} 
A key limitation of the factor-wise loss is that it enforces consistency only locally, at the level of individual factors, without guaranteeing global coherence across the full assignment space. In particular, marginals predicted for different factors may not correspond to any valid joint distribution. While LBP partially mitigates this by promoting local consistency during message passing, the lack of a true likelihood objective means the model may overfit to local patterns that do not translate into improved global solutions. Addressing this limitation through likelihood approximations or structured regularization remains an open direction. Another limitation is that many combinatorial optimization instances may have multiple optimal or near-optimal solutions. Since the supervised target provides only one reference solution $\YVEC_i$, the loss may penalize local configurations that belong to alternative equally good solutions. Thus, the objective trains the model to reproduce the provided reference solution rather than to represent the full set of optimal assignments.

\input{resources/tikz/chapter_3/model_design_efcanonicalparam}

\section{Experiments}
\label{sec:chapter_3_experiments}
In this section, we evaluate four claims about learned MRF inference for combinatorial optimization. First, learned MRFs improve feasible-solution decoding with both EGG and GD, compared with direct neural prediction and non-learned canonical MRF baselines. Second, GD improves solution construction because evidence clamping changes the remaining marginals during the dive. Third, the learned non-unary factor potentials encode local feasibility information by separating feasible from infeasible configurations. Fourth, the benefit of additional inference saturates after a finite budget, so inference cost must be balanced against solution quality.

We evaluate these claims on two families of synthetic combinatorial optimization problems: MIS and SC. We first describe the experimental protocol, including dataset generation, feature extraction, model hyperparameters, training objective, optimizer, number of epochs, and hardware used for training and evaluation. We then introduce the predictors and decoding strategies considered in the comparison: non-learned canonical MRF baselines, a direct GNN logit predictor, and learned MRF models using either LBP or NMF inference. Finally, we present the main results, analyze GD and learned factor potentials, study the inference-budget trade-off, and evaluate the fix-and-solve variant.

\subsection{Experimental protocol}
This subsection describes the common experimental protocol used to compare the different prediction and inference methods. We first specify the families of predictors considered in the experiments, including the non-learned probabilistic baselines, the direct GNN logit predictor, and the learned MRF-based models. We then describe the shared training setup, the approximate inference budget, the decoding strategies used to convert scores or marginals into feasible solutions, and the metrics used to evaluate solution quality and runtime. For GD, we also evaluate cheaper update schedules that rerun inference only every $k$ decoding steps.

\paragraph{Compared methods}
We compare five prediction methods, summarized in Table~\ref{tab:chapter_3_compared_methods}. All method names are defined before the table to make the comparison unambiguous. \emph{CMRF-LBP} and \emph{CMRF-NMF} are non-learned canonical-MRF baselines: their factor parameters are fixed by the feasibility-cost parameterization in Section~\ref{subsec:canonical-feasibility-cost-param}, and LBP or NMF is then used only as an approximate inference procedure. \emph{DGN} is the direct GNN baseline: it uses the same neural backbone as the learned MRF models but outputs one logit per decision variable and does not include an MRF inference layer. \emph{LMRF-LBP} and \emph{LMRF-NMF} are learned-MRF predictors: a GNN predicts over-complete factor parameters, and LBP or NMF converts those parameters into approximate variable marginals.

The table separates the quantity used by the decoder from the quantity used for training. The decoder always receives one scalar score per decision variable: a direct logit for DGN, or an approximate variable marginal for the MRF-based methods. The training signal differs across learned models: DGN is trained from variable-level labels, LMRF-LBP uses the factor-wise marginal surrogate from Section~\ref{sec:chapter_3_training_objective}, and LMRF-NMF uses the variable-wise marginal surrogate. The CMRF variants contain no learned parameters, so they do not have a supervised training objective.

\begin{table}[t]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{lccccc}
\toprule
\textbf{Method}    & \textbf{Trained} & \textbf{MRF parameters}    & \textbf{Inference} & \textbf{Loss} & \textbf{Decoder input} \\
\midrule
DGN      & Yes     & --              & -- & $\BCELOSS$ & Logits \\
CMRF-LBP & No      & $\CANEPARAM(\PROB)$     & LBP & -- & $\MARGINAL_i$ \\
CMRF-NMF & No      & $\CANEPARAM(\PROB)$     & NMF & -- & $\MARGINAL_i$ \\
LMRF-LBP & Yes     & $\PREEPARAM(\PROB;\NNPARAM)$     & LBP & $\LOSSF^{\mathrm{LBP}}$ & $\MARGINAL_i$ \\
LMRF-NMF & Yes     & $\PREEPARAM(\PROB;\NNPARAM)$     & NMF & $\LOSSF^{\mathrm{NMF}}$ & $\MARGINAL_i$ \\
\bottomrule
\end{tabular}}
\caption{Summary of the prediction methods compared in the experiments. The column ``Trained'' indicates whether the method contains trainable neural parameters; CMRF parameters depend on the problem instance but are fixed by the canonical construction.}
\label{tab:chapter_3_compared_methods}
\end{table}

Thus, CMRF-LBP and CMRF-NMF test whether the hand-designed canonical MRF parameterization is useful when paired with approximate inference, while LMRF-LBP and LMRF-NMF test whether learned factor parameters produce more useful approximate marginals under the same inference budget. DGN isolates the effect of removing the MRF layer entirely.

\paragraph{Training setup} 
All learned models are trained using the supervised objective introduced in Section~\ref{sec:chapter_3_training_objective}. DGN is trained from variable-level labels, whereas the learned MRF models are trained through the marginals returned by the chosen inference procedure: factor marginals for LMRF-LBP and variable marginals for LMRF-NMF.

All learned models use the same neural backbone, a four-layer GCN \citep{kipf_semi-supervised_2017} with hidden dimension $\EMBDIM=64$. This shared architecture keeps the comparison computationally tractable and isolates the effect of the prediction head and inference layer. The direct predictor outputs one logit $\SCORE_i$ per decision variable, while the learned MRF models output factor parameters $\EPARAM_{\FACTOR}$ that are passed to either LBP or NMF inference.

Models are optimized with AdamW~\citep{loshchilov_decoupled_2018-1} using a learning rate of $10^{-3}$ across all datasets. Approximate inference is run for at most $10$ iterations, and the LBP damping coefficient is set to $0.5$. Each dataset is split into training, validation, and test sets using an $80/10/10\%$ split. Results are reported over three random seeds, $\{0,1,2\}$.

\paragraph{Inference and decoding}
At test time, each method produces one score per decision variable. DGN outputs logits directly, while CMRF-LBP, CMRF-NMF, LMRF-LBP, and LMRF-NMF obtain variable marginals by running approximate inference on the corresponding MRF. Inference is capped at $10$ iterations for both LBP and NMF, with LBP message damping coefficient set to $0.5$. Since the factor graphs are loopy, LBP beliefs are treated as approximate marginals.

Scores or marginals are converted into feasible solutions using two decoders. EGG is a static decoder: it computes predictions once and fixes variables in decreasing order of confidence, measured by low entropy for marginals or by the sigmoid confidence of the DGN logit. GD (Algorithm~\ref{alg:bp_guided_decimation}) is adaptive: after each assignment, the model is conditioned on the current partial solution and inference is rerun before selecting the next variable. EGG therefore evaluates the quality of the initial predictions, whereas GD evaluates whether the inferred distribution remains useful under iterative partial assignments.

\paragraph{Evaluation metrics}
Assuming that all problems are given in canonical form, we evaluate solution quality using the relative gap
\begin{equation}
    \mathrm{gap}(\hat{\YVEC})
    =
    \frac{
        z(\YVEC^{\mathrm{ref}}) - z(\hat{\YVEC})
    }{
        |z(\YVEC^{\mathrm{ref}})|
    },
\end{equation}

where $z(x)$ is the objective value of the solution returned by the method and $z(\YVEC^{\mathrm{ref}})$ is the objective value of the reference solution provided with the dataset. Since the dataset labels are produced by external solvers or heuristics, this quantity should be interpreted as a gap with respect to the available reference solution, not necessarily as a gap to the true optimum. Lower values indicate better solutions, and a zero gap means that the method matches the reference objective value.

We also report end-to-end solve time at test time. This includes all steps required to produce a feasible solution from an input instance, including preprocessing, neural prediction, approximate inference, and decoding. For runtime-split figures, we decompose this total time into its main components in order to distinguish the cost of prediction, inference, and solution construction.

\subsection{Datasets and instance generation}
\label{sec:datasets}

We evaluate this framework on two families of synthetic binary optimization problems: MIS and SC. These problems were chosen because they admit compact LP formulations, have natural greedy construction policies, and induce factor graphs whose factor arity can be controlled through the instance-generation process. This makes them suitable for evaluating both the quality of learned predictions and the behavior of MRF-based inference
under different constraint structures. We restrict this chapter to the small and medium regimes; larger instances
are considered in Chapter~\ref{chapter:5_feature-to-factor_mappings_for_approximate_mrf_inference}.

\paragraph{Maximum Independent Set}
MIS instances are generated from Erdős--Rényi random graphs with edge probability $15\%$. We consider two size regimes: $\NVAR = 100$ and $\NVAR \sim \mathcal{U}[200,300]$. For graphs with more than $100$ vertices, node degrees are bounded by $15$ in order to control the density of the resulting factor graph. Reference labels are produced by solving the instances with the KaMIS solver~\citep{lamm_finding_2017}.

\paragraph{Set Cover}
SC instances are generated using a balanced randomized procedure. For a given number of items $n$, sets $m$, and coverage parameter $k$, we sample an item--set incidence matrix and enforce two validity conditions: no set is empty, and each item is covered by exactly $k$ sets. We consider two size regimes: $n=50$, $m=50$, $k=5$, and $n \sim \mathcal{U}[120,180]$, $m \sim \mathcal{U}[150,250]$, $k=5$. Reference labels are obtained using the SCIP solver~\citep{achterberg_scip_2009}.

\input{resources/tab/chapter_3/dataset_identifiers.tex}

\paragraph{Features}
The input representation uses $54$ features, divided into two categories. The first $10$ are extracted from the continuous relaxation of the problem, including LP solution values and related simplex information. The remaining $44$ are graph-based features computed from the variable--constraint graph. Together, these features provide both optimization-aware information from the relaxation and structural information from the underlying factor graph. The full list of features is given in Table~\ref{tab:dataset_features}.

\input{resources/tab/chapter_3/dataset_features.tex}

\subsection{Main results: learned inference improves decoding}
\label{subsec:chapter_3_main_results}

Learned MRF inference improves both static EGG decoding and adaptive GD decoding across the reported SC and MIS benchmarks. Table~\ref{tab:mrf_nair_inference_results} reports the average relative optimality gap and runtime for the evaluated inference and decoding variants. The same results are visualized in Figure~\ref{fig:mrf_nair_inference_grouped_bar_optimality}, which compares optimality gaps, and Figure~\ref{fig:mrf_nair_inference_stacked_grouped_bar_runtime}, which decomposes the runtime. Across the evaluated settings, LMRF-LBP provides the strongest overall decoding signal: it improves over the corresponding canonical LBP baseline and over the direct neural baseline in the SC and MIS-S comparisons.

A useful way to interpret the difference between the canonical MRFs, CMRF-LBP and CMRF-NMF, and the learned MRFs, LMRF-LBP and LMRF-NMF, is through the approximate nature of the inference procedure used at test time. Since the graph instances contain loops, the marginals produced by loopy belief propagation or mean-field inference are not exact marginals of the underlying MRF. In addition, inference is run for a bounded number of iterations, so the algorithm may stop before full convergence. Therefore, the decoder receives finite-iteration approximate marginals rather than exact probabilistic marginals.

This distinction is important because learning adapts the model parameters to the inference procedure that is actually used during decoding. In other words, the learned MRF is not only fitting the observed solutions, but is also fitting them through the approximate inference dynamics induced by LBP or NMF. This can make the learned parameters better suited to the finite-iteration approximate marginals available at test time. By contrast, the canonical parameters used in CMRF-LBP and CMRF-NMF may be more faithful to the original hand-designed MRF formulation, but they are not necessarily optimized for the bounded inference scheme used in practice. As a result, they may require more iterations, or more accurate convergence, before producing marginals that are equally useful for decoding.

\input{resources/tab/chapter_3/mrf_nair_inference_results.tex}

On the SC instances, learned inference leads to a clear reduction in optimality gap. For SC-S, LMRF-LBP reduces the EGG gap to $3.748\%$, compared with $13.239\%$ for CMRF-LBP and $7.800\%$ for DGN. On SC-M, LMRF-LBP reaches an EGG gap of $5.042\%$, compared with $15.490\%$ for CMRF-LBP and $10.320\%$ for DGN. Thus, even before adaptive decimation is used, learned MRF inference produces variable marginals that are more useful for greedy feasible-solution construction than either fixed canonical MRF parameters or direct independent variable scores.

The improvement is even more pronounced when using GD decoding. On SC-S, LMRF-LBP reaches a relative gap of $1.003\%$, while on SC-M it reaches $0.274\%$. These are the strongest SC results in Table~\ref{tab:mrf_nair_inference_results}. This indicates that learned inference and GD decoding are complementary: the learned model provides informative marginals, while the decoder converts this signal into a high-quality feasible solution by conditioning on earlier decisions and recomputing the remaining marginals.

\input{resources/tikz/chapter_3/mrf_nair_inference_grouped_bar_optimality.tex}

The same qualitative behavior is observed on the MIS instances. The MIS datasets are more challenging overall, as reflected by the higher relative gaps across all methods. Nevertheless, LMRF-LBP remains the strongest method in the reported MIS comparisons. On MIS-S, it obtains an EGG gap of $15.575\%$, improving over CMRF-LBP at $18.192\%$, CMRF-NMF at $19.177\%$, and DGN at $22.105\%$. With GD decoding, the LMRF-LBP gap is further reduced to $3.488\%$. On MIS-M, LMRF-LBP also improves over the corresponding canonical baselines, reaching $19.443\%$ with EGG decoding and $3.082\%$ with GD decoding. The reduction from EGG to GD is particularly important on MIS because each selected vertex can constrain many neighboring variables; recomputing beliefs after each decision helps the decoder react to these local exclusion effects.

Runtime results show the expected trade-off between decoding strength and computational cost. EGG decoding is faster because it uses one set of predictions throughout the dive, while GD decoding often achieves substantially smaller optimality gaps by repeatedly conditioning on the current partial assignment and rerunning inference. This trade-off is visible in Table~\ref{tab:mrf_nair_inference_results} and is made clearer by the runtime decomposition in Figure~\ref{fig:mrf_nair_inference_stacked_grouped_bar_runtime}. For example, on SC-M, LMRF-LBP improves from $5.042\%$ with EGG decoding to $0.274\%$ with GD decoding, while runtime increases from $1.156$s to $3.839$s. On MIS-M, the same move reduces the gap from $19.443\%$ to $3.082\%$, while runtime increases from $4.146$s to $4.885$s. EGG therefore remains useful when fast decoding is required, whereas GD is preferable when solution quality is the primary objective.

\input{resources/tikz/chapter_3/mrf_nair_inference_stacked_grouped_bar_runtime.tex}

Overall, these results support the central claim of this section: learned inference improves decoding. Across the reported benchmarks, LMRF-LBP produces consistently better feasible solutions than the corresponding canonical inference baselines and the direct neural decoder. The comparison between CMRF-LBP/CMRF-NMF and LMRF-LBP/LMRF-NMF also suggests that learning is especially beneficial when inference is approximate and iteration-limited, since the learned parameters can adapt to the marginals produced by the actual inference procedure used at test time. The gains are robust across both problem families, and they become especially strong when learned inference is paired with GD decoding.

\subsection{Analysis of guided decimation}

The GD analysis shows that evidence changes the remaining marginals, rather than merely selecting variables from a fixed initial ranking. Figure~\ref{fig:bpgd_delta_marginals} tracks how marginal probabilities change throughout the decimation process after evidence is clamped. This diagnostic directly tests whether GD uses inference adaptively during decoding.

\input{resources/tikz/chapter_3/bpgd_delta_marginals.tex}

Figure~\ref{fig:bpgd_delta_marginals} reports the average change in marginals before and after each clamping step, with $\Delta$ defined as the marginal before the update minus the marginal after the update. Positive values therefore indicate that the marginal decreases after evidence is clamped. Since GD repeatedly fixes variables and then updates the remaining beliefs under the new partial assignment, this quantity measures how strongly each clamping step affects the residual inference problem.

The figure shows that the curve for variables with ground-truth label $1$ remains consistently above the curve for variables with ground-truth label $0$ during most of the decimation process. This suggests that clamping decisions have a stronger effect on variables that belong to the reference solution. In other words, the variables that are ultimately selected are more sensitive to the evolving partial assignment, reflecting their stronger coupling to the feasibility and objective structure of the instance.

The magnitude of the marginal change also varies over the decimation trajectory. During the early and middle iterations, the changes are relatively stable, indicating that each clamping step produces a consistent update to the remaining beliefs. Toward the later iterations, the marginal changes increase before eventually dropping close to zero. This pattern is expected: as more variables are fixed, the remaining unfixed variables become more constrained, so each additional decision can have a larger effect on the residual problem. At the final iterations, only a small number of variables remain undecided, and the average marginal change necessarily vanishes.

This behavior highlights the difference between GD and EGG. EGG makes all decisions from a single set of scores or marginal estimates. GD, in contrast, repeatedly revises the marginal estimates after each partial assignment. The decoder therefore does not simply follow a fixed ordering induced by the initial marginals; it uses the MRF inference procedure to propagate the consequences of previous decisions to the remaining variables.

Overall, Figure~\ref{fig:bpgd_delta_marginals} supports the interpretation of GD as an inference-aware decoder. By alternating between variable fixing and marginal recomputation, GD allows the model to revise its beliefs throughout the construction of the solution. This explains why GD can substantially reduce the optimality gap compared with EGG: each decision is made using marginals that are conditioned on the current partial assignment, making the final solution more consistent with the constraints.

\subsection{Learning constraint-aware factor potentials}

The learned non-unary factor potentials separate feasible from infeasible local configurations, which explains why learned MRF inference can provide a stronger decoding signal than independent variable scoring. Figure~\ref{fig:inference_theta_distribution} shows the distribution of predicted interaction parameters, separated according to whether the corresponding local assignment is feasible or infeasible. This diagnostic allows us to inspect whether the learned MRFs encode constraint information directly in their factor potentials, rather than relying only on unary variable scores.

\input{resources/tikz/chapter_3/inference_theta_distribution.tex}

Figure~\ref{fig:inference_theta_distribution} shows that the learned MRFs assign substantially different interaction parameters to feasible and infeasible local configurations. For both LMRF-LBP and LMRF-NMF, infeasible configurations are shifted toward a separate region of the predicted $\EPARAM$ space, while feasible configurations occupy a different range. This indicates that the learned MRFs do not only learn unary preferences over variables, but also learn non-unary signals that distinguish constraint-compatible from constraint-violating local assignments.

For set covering, this separation is particularly clear. Infeasible local assignments receive strongly negative predicted interaction parameters, whereas feasible assignments are concentrated closer to the central and positive range. This suggests that the learned factor potentials encode the structure of the covering constraints: local assignments that violate feasibility are penalized, while assignments compatible with the constraints are assigned more favorable interaction values. As a result, the approximate marginals produced by inference are informed not only by individual variable scores, but also by local feasibility patterns.

The MIS instances show the same qualitative behavior, but with a sharper and more polarized structure. This is consistent with the nature of MIS constraints: selecting adjacent vertices creates a hard local incompatibility. The learned MRFs represent this incompatibility through strongly separated interaction values for feasible and infeasible assignments. The resulting parameter separation provides an explicit mechanism for propagating constraint information during inference.

Overall, Figure~\ref{fig:inference_theta_distribution} supports the interpretation that the learned MRFs are constraint-aware. The learned non-unary factors encode local feasibility information, which helps explain why LMRF-LBP and LMRF-NMF produce more useful marginals for decoding than models that rely only on fixed canonical factor parameters or direct neural scores.

\subsection{Ablation study: inference budget and decimation frequency}

The inference-budget ablation shows that most of the benefit is obtained after a finite number of inference iterations, after which additional iterations give diminishing returns. Figure~\ref{fig:inference_freq_max_iter_ablation_pareto} studies the runtime--quality trade-off induced by two GD hyperparameters: the reinference interval and the maximum number of inference iterations. Both hyperparameters are swept over $\{1,2,4,8,16,32,64\}$. The left column varies the reinference interval with the inference budget fixed to $10$ iterations, while the right column varies the inference budget with the reinference interval fixed to $1$. Each point reports the average runtime and optimality gap for one setting, with lower and further-left points corresponding to better Pareto performance.

The first trend is that reinference frequency has a strong effect on solution quality. When reinference is performed more frequently, GD has more opportunities to update the remaining marginals after each partial assignment. This generally improves the final optimality gap, but also increases runtime because inference must be repeated more often during decimation. Conversely, increasing the reinference interval reduces runtime, since several variables are fixed before the next inference call, but the decoder then relies for longer on stale marginal information. This can hurt solution quality, especially on the MIS instances where local feasibility constraints are highly sensitive to previous selections.

This behavior is clear for LMRF-LBP with GD. On MIS-S, using the most frequent reinference setting gives the lowest gap, while larger intervals reduce runtime but produce substantially worse gaps. The same pattern appears on MIS-M, where the gap increases steadily as reinference becomes less frequent. On SC-S and SC-M, the degradation is less severe for moderate reinference intervals, suggesting that the SC instances tolerate some reuse of marginals. However, very infrequent reinference still worsens solution quality, showing that the benefit of GD comes from repeatedly adapting the marginals to the current partially fixed solution.

\input{resources/tikz/chapter_3/inference_freq_max_iter_ablation_pareto.tex}

The second trend is that increasing the maximum number of inference iterations does not always lead to proportional improvements. In the right column of Figure~\ref{fig:inference_freq_max_iter_ablation_pareto}, larger iteration budgets consistently increase runtime, but the resulting gap often improves only up to a moderate number of iterations and then saturates. For LMRF-LBP on MIS-S, the gap improves as the budget increases from very small values, but the gains become limited at larger budgets. A similar saturation effect is visible on MIS-M and SC-M, where moving to very large iteration counts adds substantial runtime without a comparable reduction in optimality gap.

This saturation is consistent with the approximate-inference interpretation discussed earlier. Since the models are evaluated on loopy graphs and inference is bounded, the goal is not necessarily to recover fully converged exact marginals. Instead, GD needs marginals that are sufficiently informative for the next decimation decisions. After a moderate number of iterations, additional message-passing updates may refine the beliefs, but these refinements do not always change the discrete decisions made by the decoder. As a result, the best runtime--quality trade-off is often achieved by using a moderate inference budget rather than maximizing the number of iterations.

The comparison between LMRF-LBP and LMRF-NMF also shows that the two approximate inference schemes respond differently to the ablation. LMRF-NMF is generally faster, reflecting the lower cost of mean-field-style updates, but it often requires a careful choice of reinference interval and iteration budget to reach competitive gaps. LMRF-LBP is more expensive per inference call, but it tends to provide stronger marginal updates, especially on the more constrained MIS instances. This explains why LMRF-LBP often occupies the better-quality region of the Pareto frontier, while LMRF-NMF can be attractive in lower-runtime regimes.

Overall, the ablation confirms that the effectiveness of GD depends on balancing two forms of inference budget: how often inference is repeated during decimation, and how much message passing is performed at each reinference step. Frequent reinference improves adaptivity but increases runtime, while larger iteration budgets improve the quality of each marginal update but quickly reach diminishing returns. The best configurations therefore lie in the intermediate regime, where the decoder receives sufficiently fresh and informative marginals without spending excessive time on over-refining approximate beliefs.

\subsection{Fix-and-solve with learned predictions}

Fix-and-solve is most useful at moderate fixing ratios, where learned predictions reduce the solver workload without overconstraining the residual problem. We evaluate this by using learned predictions to accelerate an exact solver through a fix-and-solve procedure. The evaluation metric is the relative optimality gap with respect to a reference optimum. For each instance, the reference solution is obtained by solving the original problem to optimality with Gurobi~\cite{gurobi}. In the fix-and-solve experiments, however, the residual problem is solved with SCIP under a restrictive time limit of $300$ seconds. Thus, the reported gaps measure the quality of the SCIP solution obtained after partial fixing, relative to the optimal objective value computed independently by Gurobi.

In this setting, the learned model is not used to produce the final solution directly. Instead, it is used to decide a subset of variables whose values are fixed before calling SCIP on the remaining reduced problem. More precisely, for a target fixed-variable proportion $\rho$, the decoder is run until $\rho$ of the variables have been assigned. The selected variables and their predicted values are then clamped in the original combinatorial problem, and SCIP is used to solve the residual instance over the remaining unfixed variables within the $300$ second time limit. This protocol is applied in the same way for both EGG and GD decoding. With EGG, variables are fixed according to the ordering induced by the initial scores or marginals until the desired proportion is reached. With GD, the decimation process is stopped once the target proportion has been fixed; inference and clamping are performed as in the standard GD decoder, but the remaining variables are left to SCIP rather than being decoded by GD. Thus, the reported solution is always obtained by SCIP on a partially fixed problem, and the learned model only determines which variables are fixed before the exact solve.

\input{resources/tikz/chapter_3/inference_test_nair_gap_bar.tex}

Figure~\ref{fig:inference_test_nair_gap_bar} reports the resulting relative optimality gaps for different fixed-variable proportions. Since the reference value is the Gurobi optimum of the original instance, these gaps quantify how much solution quality is lost when SCIP is asked to solve only the reduced problem induced by the learned partial assignment. The main trend is that fixing more variables generally increases the final gap. This is expected: as $\rho$ grows, the reduced SCIP problem becomes smaller, but the solver has less freedom to correct mistakes made by the learned decoder. At low fixing proportions, many variables remain free, so SCIP can still repair uncertain predictions. At high fixing proportions, the quality of the final solution depends much more strongly on the correctness of the fixed assignments.

In the set-covering experiments, learned MRF predictions are effective for fix-and-solve. LMRF-LBP with EGG and GD both achieve small gaps at moderate fixing proportions, indicating that a substantial fraction of variables can be fixed while still leaving SCIP enough flexibility to recover a high-quality solution. For example, on SC-S and SC-M, the gap remains low when fixing $50\%$ or $60\%$ of the variables, and the degradation is gradual as the fixed-variable proportion increases. This suggests that the learned marginals provide a reliable ranking of variables for the set-covering instances.

The behavior on MIS is more delicate. The gaps increase more rapidly as the fixed-variable proportion grows, especially for GD-based fixing. This reflects the harder local structure of MIS: fixing a vertex incorrectly can immediately constrain several neighboring vertices and may remove feasible high-quality completions from the reduced problem. In this setting, the fix-and-solve procedure is more sensitive to early mistakes, because SCIP can only optimize over the variables that remain unfixed. The MIS results therefore show that aggressive fixing is riskier on problems with strong exclusion constraints.

An important distinction from the complete-decoding experiments in Section~\ref{subsec:chapter_3_main_results} is that GD is not necessarily always the best choice for fix-and-solve. When GD is used as a complete decoder, its repeated inference updates help construct a full feasible solution. In fix-and-solve, however, the goal is different: the decoder should fix only the most reliable variables and leave the uncertain part of the instance to SCIP. For MIS in particular, GD can make confident sequential commitments that are useful for full decoding but too restrictive when many of those commitments are imposed as hard constraints before the solver is called. EGG can therefore be competitive, or even preferable, because it provides a simpler static fixing rule that may be less aggressive in propagating early decisions.

\input{resources/tikz/chapter_3/inference_test_nair_runtime_stacked_bar.tex}

Figure~\ref{fig:inference_test_nair_runtime_stacked_bar} shows the corresponding runtime decomposition. The total runtime is split into preprocessing, model forward pass, decoding-based fixing, and the final reduced SCIP solve. The dominant trend is that the reduced-solver time decreases as the fixed-variable proportion increases. This confirms the intended effect of fix-and-solve: by clamping a larger part of the problem, the exact solver is called on a smaller residual instance and therefore requires less time within the $300$ second time limit.

However, the runtime reduction comes with the gap increase observed in Figure~\ref{fig:inference_test_nair_gap_bar}. The fixed-variable proportion therefore controls the main quality--runtime trade-off. Small values of $\rho$ are conservative: they preserve solution quality but leave a larger problem for SCIP. Large values of $\rho$ are aggressive: they accelerate the reduced solve but can lock in incorrect assignments. Intermediate fixing proportions provide the most useful regime, since they reduce the SCIP workload while still allowing the solver to correct the uncertain part of the solution.

The runtime decomposition also highlights the different costs of EGG and GD. EGG has a relatively small decoding cost because the variables are selected from a single set of predictions. GD is more expensive because it repeatedly clamps variables and updates the remaining marginals before reaching the target fixed-variable proportion. This additional cost can be worthwhile when the resulting fixed assignments are more reliable, but it is not automatically beneficial in the fix-and-solve setting. The best method is therefore determined by the combined effect of prediction quality, decoding cost, and reduced-solver time.

\input{resources/tikz/chapter_3/inference_test_nair_pareto.tex}

Figure~\ref{fig:inference_test_nair_pareto} summarizes this trade-off as a Pareto plot, where each point corresponds to one fixed-variable proportion. Points closer to the lower-left corner are preferable because they combine lower runtime with lower optimality gap. The Pareto view makes clear that there is no single universally optimal fixing proportion. Instead, each value of $\rho$ represents a different operating point: lower fixing proportions favor accuracy, while higher fixing proportions favor speed.

Across the SC datasets, LMRF-LBP provides strong performance, especially at moderate fixing proportions. This indicates that the learned MRF predictions can safely reduce the problem size while preserving much of the solution quality. LMRF-NMF is often faster, reflecting the lower cost of mean-field-style inference, but may require a more conservative fixing proportion to avoid larger gaps. DGN is competitive in some fast regimes, but the learned MRF variants generally provide a better quality signal for deciding which variables to fix.

On the MIS datasets, the Pareto frontier is more constrained. The gap rises quickly when too many variables are fixed, so the useful region is concentrated at lower or moderate fixing proportions. This again reflects the fact that MIS feasibility is highly sensitive to local assignment errors. For these instances, fix-and-solve is most effective when the learned decoder is used conservatively: it should fix only the variables for which the prediction is reliable, while leaving enough freedom for SCIP to resolve the remaining combinatorial structure.

Overall, the fix-and-solve experiments show that learned predictions can be used not only as standalone decoders, but also as a mechanism for reducing exact optimization problems. The key requirement is to treat the learned decoder as a partial assignment generator rather than as a replacement for the solver. By fixing a controlled fraction of variables and then solving the residual problem with SCIP, the method provides a tunable trade-off between runtime and solution quality. The best results are obtained in the intermediate regime, where the learned model fixes enough variables to simplify the instance, but not so many that the reduced problem becomes overconstrained by prediction errors.

\section{Discussion and limitations}

The experiments in this chapter show that learned MRF inference provides a more useful decoding signal than both independent neural predictions and canonical, non-learned MRF parameters. This is most visible for LMRF-LBP, which consistently gives the strongest reported decoding results across the SC and MIS benchmarks. The improvement is especially pronounced when learned inference is paired with GD, because GD repeatedly conditions on partial assignments and recomputes the remaining marginals rather than relying on a fixed initial ranking.

These gains should be interpreted in light of the approximate inference procedure used at test time. The factor graphs are loopy, and both LBP and NMF are run for a bounded number of iterations. The decoder therefore receives finite-iteration approximate beliefs, not exact marginals. Learning is beneficial because it adapts the MRF parameters to this practical inference regime. The learned factor-potential analysis supports this interpretation: the non-unary factors learn to separate feasible from infeasible local configurations, so the resulting beliefs encode constraint information that is absent from purely factorized neural scores.

The ablation and fix-and-solve experiments clarify the main trade-offs. GD improves solution quality by using inference adaptively, but frequent reinference and larger iteration budgets increase runtime, with diminishing returns after a moderate budget. Similarly, fix-and-solve is useful when the learned model fixes a controlled fraction of variables, especially on SC instances, but aggressive fixing can overconstrain the residual problem, particularly for MIS. Thus, the learned MRF should be viewed as a structured source of partial information rather than as a replacement for exact optimization.

Two limitations remain. First, the explicit over-complete factor-potential parameterization does not scale to large factors or large local state spaces. It is feasible for the MIS and SC formulations studied here because the relevant factors remain small, but it would become impractical for problems whose natural constraints involve many variables. This motivates the more scalable parameterizations studied later in Chapter~\ref{chapter:5_feature-to-factor_mappings_for_approximate_mrf_inference}.

Second, the strongest decoding results often rely on GD with frequent reinference. Each reinference step recomputes approximate marginals under a new partial assignment, and this repeated inference can dominate runtime. The ablation study shows that this cost can be controlled by reducing the reinference frequency or limiting the number of inference iterations, but doing so weakens the adaptivity that makes GD effective. The method therefore provides a tunable quality--runtime trade-off rather than a uniformly faster replacement for simpler decoders.

\section{Conclusion}

This chapter developed learned MRF inference as a structured alternative to fully factorized neural decoding for combinatorial optimization. By replacing independent variable scores with learned factor potentials and approximate MRF inference, the model produces beliefs that incorporate both objective information and local constraint structure. The experiments show that this structure is useful in practice: learned MRFs, especially LMRF-LBP combined with guided decimation, provide stronger decoding signals than direct neural predictions and canonical non-learned MRF baselines.

The main conclusion is that learned MRFs are most effective not as exact optimizers, but as inference-aware predictors that guide downstream decision procedures. They can improve one-shot decoding, adaptive guided decimation, and hybrid fix-and-solve workflows when used with an appropriate inference budget and a conservative fixing strategy. At the same time, the explicit over-complete parameterization studied here limits scalability, which motivates the feature-to-factor mapping framework of Chapter~\ref{chapter:5_feature-to-factor_mappings_for_approximate_mrf_inference}.
